


Cocaine Addiction
========================================================


Two different methods to read in the data are illustrated next.

```{r Cocaine}
Outcome <- c(rep("relapse", 10), rep("no relapse", 14), rep("relapse", 18), rep("no relapse", 6))
Drug <- c(rep("Desipramine", 24), rep("Lithium", 24))
Cocaine <- data.frame(Outcome, Drug)
head(Cocaine)
ANS <- with(data = Cocaine,
            tapply(Outcome == "relapse", Drug, mean)
            )
ANS
obsDiffProp <- ANS[1] - ANS[2]
names(obsDiffProp) <- NULL
obsDiffProp
```

```{r drug2}
Outcome2 <- c(rep(1, 10), rep(0, 14), rep(1, 18), rep(0, 6))
Drug2 <- c(rep("Desipramine", 24), rep("Lithium", 24))
Cocaine2 <- data.frame(Outcome2, Drug2)
head(Cocaine2)
ANS2 <- with(data = Cocaine2,
     tapply(Outcome2, Drug2, mean)
            )
ANS2
obsDiffProp2 <- ANS2[1] - ANS2[2]
names(obsDiffProp2) <- NULL
obsDiffProp2
```

Regardless of the method used to read in the data, the observed difference in proportion of addicts who relapse using Desipramine and Lithium is `r round(obsDiffProp, 4)`. 


Repeating the Experiment
========================

The following code can be used when reading the data into a data frame using the second approach.

```{r coke1rep}
N <- 10^4 -1
DiffProp <- numeric(N)
set.seed(4)
for (i in 1:N) {
# sample of size 24 # from the total 48 addicts
index <- sample(48, size = 24, replace = FALSE)
DiffProp[i] <- mean(Cocaine2$Outcome2[index]) - mean(Cocaine2$Outcome2[-index])
}
hist(DiffProp, col = "blue", breaks = "Scott", xlab = "", main = "")
abline(v = obsDiffProp2, col = "red")
pvalue1 <- (sum(DiffProp <= obsDiffProp2) + 1)/(N + 1)
pvalue1
# ggplot2 approach
DF <- data.frame(x = DiffProp)
p <- ggplot(data = DF) + 
  geom_density(aes(x = x, y = ..density..), fill = 'pink', alpha = 0.4) +
  theme_bw()
p
x.dens <- density(DiffProp)
df.dens <- data.frame(x = x.dens$x, y = x.dens$y)
p + geom_area(data = subset(df.dens, x <= obsDiffProp & x >= min(DF$x)), aes(x = x, y = y), fill = 'blue', alpha = 0.4) + 
  labs(x = expression(hat(p)[Desipramine] - hat(p)[Lithium]), y = '', title = "Randomization Distribution")
```

The following code can be used when reading the data into a data frame using the first approach.

```{r coke2rep}
N <- 10^4 -1
DiffProp <- numeric(N)
set.seed(4)
for (i in 1:N) {
# sample of size 24 # from the total 48 addicts
index <- sample(48, size = 24, replace = FALSE)
DiffProp[i] <- mean(Cocaine$Outcome[index]=="relapse") - mean(Cocaine$Outcome[-index]=="relapse")
}
hist(DiffProp, col = "blue", breaks = "Scott", xlab = "", main = "")
abline(v = obsDiffProp, col = "red")
pvalue <- (sum(DiffProp <= obsDiffProp) + 1)/(N + 1)
pvalue
# ggplot2 approach
DF <- data.frame(x = DiffProp)
p <- ggplot(data = DF) + 
  geom_density(aes(x = x, y = ..density..), fill = 'pink', alpha = 0.4) + 
  theme_bw()
p
x.dens <- density(DiffProp)
df.dens <- data.frame(x = x.dens$x, y = x.dens$y)
p + geom_area(data = subset(df.dens, x <= obsDiffProp & x >= min(DF$x)), aes(x = x, y = y), fill = 'blue', alpha = 0.4) + 
  labs(x = expression(hat(p)[Desipramine] - hat(p)[Lithium]), y = '', title = "Randomization Distribution")
```

Recall that if the true _p_-value is 0.02, then we can expect the standard deviation of our simulated _p_-value to be approximately $\sqrt{p\times(1-p)/N}$, or approximately `r round((.02*.98/N)^.5, 4)`.

Mathematical Symbols
=======================

There are many online references for writing in $\LaTeX$.  Here is a [link to a pdf](http://mirrors.ibiblio.org/CTAN/info/examples/Math_into_LaTeX-4/SymbolTables.pdf) that you may find useful.

Try writing a few equations with $\LaTeX$.  Once you get the hang of $\LaTeX$, you will $\heartsuit$ it!  

$$\bar{y} = \sum_{i=1}^{n} \frac{y_i}{n}$$
