---
output: pdf_document
---



Exploratory Data Analysis
========================================================

## Reading *.csv Data
```{r READIN}
site <- "http://www1.appstate.edu/~arnholta/Data/FlightDelays.csv"
FlightDelays <- read.csv(file=url(site))
head(FlightDelays) # shows first 6 rows of data frame
str(FlightDelays)  # shows structure of data frame
levels(FlightDelays$Month)
FlightDelays$Month <- factor(FlightDelays$Month, levels = c("May", "June"))
levels(FlightDelays$Month)
```

## Creating Tables
```{r TABLES}
table(FlightDelays$Carrier)
xtabs(~Carrier, data = FlightDelays)
```

## Creating Barplots
```{r Barplots, tidy = FALSE}
barplot(table(FlightDelays$Carrier))
require(ggplot2)
ggplot(data = FlightDelays, aes(x = Carrier)) + 
  geom_bar()
ggplot(data = FlightDelays, aes(x = Carrier, fill= Month)) + 
  geom_bar()
ggplot(data = FlightDelays, aes(x = Carrier, fill= Month)) + 
  geom_bar() + 
  guides(fill = guide_legend(reverse = TRUE))
ggplot(data = FlightDelays, aes(x = Carrier, fill= Month)) + 
  geom_bar(position="dodge") + 
  guides(fill = guide_legend(reverse = TRUE))
xtabs(~ Carrier + (Delay > 30), data = FlightDelays)
addmargins(xtabs(~ Carrier + (Delay > 30), data = FlightDelays))
ggplot(data = FlightDelays, aes(x = Carrier, fill= Delayed30)) + 
  geom_bar(position="dodge")
ggplot(data = FlightDelays, aes(fill = Carrier, x= Delayed30)) + 
  geom_bar(position="dodge")
```

## Histograms of `Delay` values.

```{r HistoEDA, tidy = FALSE}
hist(FlightDelays$Delay)  # Ugly with Defaults...you change
library(lattice)
histogram(~Delay, data = FlightDelays)
histogram(~Delay, data = FlightDelays, type = "density")
histogram(~Delay, data = FlightDelays, type = "density",
          panel = function(...){
            panel.histogram(col = "peru",...)
            panel.densityplot(col = "red", lwd = 2, ...)
          }
          )
ggplot(data = FlightDelays, aes(x = Delay)) + 
  geom_histogram()
ggplot(data = FlightDelays, aes(x = Delay, y = ..density..)) + 
  geom_histogram(binwidth = 10, color = "blue") + 
  geom_density(color = "red")
ggplot(data = FlightDelays, aes(x = Delay)) + 
  geom_density(fill = "blue")
```
## Numeric Summaries

```{r NS}
summary(FlightDelays)
sd(FlightDelays$Delay)
sd(FlightDelays$Delay)^2
var(FlightDelays$Delay)
IQR(FlightDelays$Delay)
quantile(FlightDelays$Delay)
```

## Boxplots

```{r BP}
boxplot(Delay ~ Carrier, data = FlightDelays)
bwplot(Delay ~ Carrier, data = FlightDelays)
bwplot(Delay ~ Carrier|Month, data = FlightDelays, as.table = TRUE, layout = c(2, 1))
ggplot(data = FlightDelays, aes(x = Carrier, y = Delay)) + geom_boxplot()
ggplot(data = FlightDelays, aes(x = Carrier, y = Delay)) + geom_boxplot() + 
  facet_grid(. ~ Month) 
```

```{r NCBirths, warning = FALSE, tidy = FALSE}
site <- "http://www1.appstate.edu/~arnholta/Data/NCBirths2004.csv"
NCBirths <- read.csv(file=url(site))
head(NCBirths)
boxplot(Weight ~ Gender, data = NCBirths, col = c("pink", "blue"))
bwplot(Weight ~ Gender, data = NCBirths)
p <- ggplot(data = NCBirths, aes(x = Gender, y = Weight, fill = Gender))
p + geom_boxplot()
p + geom_boxplot() + 
  guides(fill = FALSE) + 
  labs( x = "Newborn Gender", y = "Weight in ounces", title = "You Put Something Here")
p + geom_boxplot() + 
  guides(fill = FALSE) + 
  labs( x = "Newborn Gender", y = "Weight in ounces", title = "You Put Something Here") + 
  scale_fill_manual(values = c('pink', 'blue'))
p + geom_boxplot() + guides(fill = FALSE) + 
  labs( x = "Newborn Gender", y = "Weight in ounces", title = "You Put Something Here") + 
  scale_fill_brewer() + theme_bw()
```

## Density Plots
```{r ND, tidy = FALSE}
curve(dnorm(x), -4, 4, ylab = "", xlab = "")
x.region <- seq(from = 1, to = 4, length.out = 200)
y.region <- dnorm(x.region)
region.x <- c(x.region[1], x.region, x.region[200])
region.y <- c(0, y.region, 0)
polygon(region.x, region.y, col = "red")
abline(h = 0, lwd = 2)
# Same now with ggplot2
p <- ggplot(data = data.frame(x = c(-4, 4)), aes(x = x))
dnorm_func <- function(x){
  y <- dnorm(x)
  y[x<1] <- NA
  return(y)
}
p1 <- p + stat_function(fun = dnorm_func, geom = 'area', fill = 'blue', alpha = 0.2) + 
  geom_hline(yintercept = 0) + 
  stat_function(fun = dnorm) 
p1 + theme_bw()
p1 + theme_bw() + 
  labs(x = '', y = '', title = expression(integral(frac(1, sqrt(2*pi))*e^{-x^2/2}*dx, 1, infinity)==0.1586553) ) # Break it down!
```

## Example 2.11 

Note this is not how `qqnorm` computes the quantiles!  The left graph of Figure 2.9 in the book is not quite correct...it does not use the data in the table...the first value 17.7 should be 21.7.

```{r QUANTILES}
x <- c(21.7, 22.6, 26.1, 28.3, 30.0, 31.2, 31.5, 33.5, 34.7, 36.0)
n <- length(x)
p <- (1:10)/(n + 1)
q <- qnorm(p)
rbind(x, p, q)
plot(q, x)
XS <- quantile(q, prob = c(.25, .75))
YS <- quantile(x, prob = c(.25, .75))
slopeA <- (YS[2] - YS[1])/(XS[2] - XS[1])
slopeB <- diff(YS)/diff(XS)
slopeA
slopeB
Intercept <- YS[1] - slopeA*XS[1]
Intercept
abline(a = Intercept, b = slopeA)
#
pc <- (1:10 - 3/8)/n
qc <- qnorm(pc)
rbind(x, pc, qc)
xs <- quantile(qc, prob = c(.25, .75))
ys <- quantile(x, prob = c(.25, .75))
slope <- diff(ys)/diff(xs)
intercept <- ys[1] - slope*xs[1]
c(intercept, slope)
```
Consider using the `R` functions `qqnorm()` and `qqline()`.

```{r QQNORM, tidy = FALSE}
qqnorm(x)
qqline(x)
# ggplot
ggplot(data = data.frame(x), aes(sample=x)) + 
  stat_qq() + 
  geom_abline(intercept = intercept, slope = slope)
```

## Empirical Cumulative Distribution Function

The _empirical cumulative distribution function_ (ecdf) is an estimate of the underlying cumulative distribution function for a sample.  The empirical cdf, denoted by $\hat{F}$, is a step function $$\hat{F}(x) = \tfrac{1}{n} (\text{number of values} \leq x),$$
where $n$ is the sample size.

```{r ECDFbase}
y <- c(3, 6, 15, 15, 17, 19, 24)
plot.ecdf(y)
set.seed(1)  # set seed for reproducibility
rxs <- rnorm(25)  
plot.ecdf(rxs, xlim = c(-4, 4))
curve(pnorm(x), col = "blue", add = TRUE, lwd = 2)
```

An alternative approach to the book's Figure 2.12 is provided using `ggplot2` after first creating Figure 2.12

```{r BEER, tidy = FALSE}
site <- "http://www1.appstate.edu/~arnholta/Data/Beerwings.csv"
Beerwings <- read.csv(file=url(site))
head(Beerwings) # shows first 6 rows of data frame
str(Beerwings)  # shows structure of data frame
beerM <- subset(Beerwings, select = Beer, subset = Gender == "M", drop = TRUE)
beerF <- subset(Beerwings, select = Beer, subset = Gender == "F", drop = TRUE)
plot.ecdf(beerM, xlab = "ounces", col = "blue", pch = 19)
plot.ecdf(beerF, col = "pink", pch = 19, add = TRUE)
abline(v = 25, lty = "dashed")
legend("topleft", legend = c("Males", "Females"), pch = 19, col = c("blue", "pink"))
# Using ggplot2 now
ggplot(data = Beerwings, aes(x = Beer, colour = Gender)) + 
  stat_ecdf() + 
  labs(x = "Beer in ounces", y ="", title = 'ECDF')
```

## Scatter Plots

```{r Bplot, tidy = FALSE}
with(data = Beerwings, plot(Hotwings, Beer, xlab = "Hot wings eaten", ylab = "Beer consumed", 
                            pch = 19, col = "blue"))
p <- ggplot(data = Beerwings, aes(x = Hotwings, y = Beer)) + 
  geom_point() + 
  labs(x = "Hot wings eaten", y = "Beer consumed in ounces")
p
p + geom_smooth()
p + geom_smooth(method = lm) + theme_bw()
```

## Integrating with R

```{r INT}
f <- function(x){(x - 1)^3*exp(-x)}
ans <- integrate(f, lower = 0, upper = Inf)$value
ans
```