

In Class Exercise (Partial Solution)
========================================================

The birth weight of a baby is of interest to health officials since many studies have shown possible links between weight and conditions later in life, such as obesity or diabetes.  The data set `Girls2004` consists of a random sample of 40 baby girls born in Alaska and 40 baby girls born in Wyoming.  These babies also had a gestation period of at least 37 weeks and were single births. 


Import the data `Girls2004` available from [http:\\www1.appstate.edu/arnholta/Data/](http:\\www1.appstate.edu/arnholta/Data/).

```{r Girls2004}
site <- "http://www1.appstate.edu/~arnholta/Data/Girls2004.csv"
Girls <- read.csv(file=url(site))
head(Girls)
```

  * Perform some exploratory data analysis and obtain summary statistics on the weights of baby girls born in Wyoming and Alaska (do separate analyses for each state).
  
```{r GirlsHist, fig.width=7, fig.height=7}
opar <- par(no.readonly = TRUE)
par(mfrow=c(2, 2))
hist(Girls$Weight[Girls$State=="AK"], col = "pink1", breaks = "Scott", xlab = "Weight in g.", main = "Alaska", freq= FALSE)
lines(density(Girls$Weight[Girls$State=="AK"]), col ="red")
qqnorm(Girls$Weight[Girls$State=="AK"], col = "pink1")
qqline(Girls$Weight[Girls$State=="AK"], col = "pink1")
hist(Girls$Weight[Girls$State=="WY"], col = "pink", breaks = "Scott", xlab = "Weight in g.", main = "Wyoming", freq= FALSE)
lines(density(Girls$Weight[Girls$State=="WY"]), col ="red")
qqnorm(Girls$Weight[Girls$State=="WY"], col = "pink")
qqline(Girls$Weight[Girls$State=="WY"], col = "pink")
par(opar)
```

Another approach is to use `lattice`.

```{r GirlsHistLattice}
require(lattice)
histogram(~Weight|State, data = Girls, layout=c(1, 2), as.table = TRUE, type = "density")
qqmath(~Weight|State, data = Girls, layout=c(1, 2), as.table = TRUE)
qqmath(~Weight, group = State, data = Girls)
```

A third approach is to use `ggplot2`. (This is my preference.)

```{r GirlsHistggplot2, message = FALSE}
require(ggplot2)
p <- ggplot(data = Girls, aes(x = Weight)) + geom_histogram(binwidth = 300, fill = "peru", color = "black") + facet_grid(State ~ .)
p + theme_bw()
p1 <- ggplot(data = Girls, aes(x = Weight)) + geom_histogram(aes(y = ..density..), binwidth = 300, fill = "blue", color = "black") + facet_grid(State ~ .) + geom_density(color = "red", size = 1)
p1 + theme_bw()
p2 <- ggplot(data = Girls, aes(sample = Weight, color = State)) + stat_qq()
p2 + theme_bw()
ggplot(data = Girls, aes(sample = Weight, shape = State, color = State)) + stat_qq() + facet_grid(State~.) + theme_bw()
```

## Graphs are fun!

```{r GirlsSummaryFun, comment = NA}
with(data = Girls, 
tapply(Weight, State, mean))
tapply(Girls$Weight, Girls$State, sd)
AK.weight <- subset(Girls, select = Weight, subset = State == "AK", drop = TRUE)
WY.weight <- subset(Girls, select = Weight, subset = State == "WY", drop = TRUE)
c(mean(AK.weight), sd(AK.weight), mean(WY.weight), sd(WY.weight))
```
> ANSWER: The mean and standard deviation of weight for baby girls born in Alaska is `r mean(AK.weight)` and `r sd(AK.weight)`, respectively.  The mean and standard deviation of weight for baby girls born in Wyoming is `r mean(WY.weight)` and `r sd(WY.weight)`, respectively. 
  
  * Bootstrap the difference in means, plot the distribution, and give summary statistics.  Obtain a 95% bootstrap percentile confidence interval and interpret this interval.
  
```{r GirlsBoot, fig.width = 7, fig.height= 3.5}
B <- 15000
mean.diff <- numeric(B)
for(i in 1:B){
  AK <- sample(AK.weight, size = sum(!is.na(AK.weight)), replace = TRUE)
  WY <- sample(WY.weight, size = sum(!is.na(WY.weight)), replace = TRUE)
  mean.diff[i] <- mean(AK) - mean(WY)
}
opar <- par(no.readonly = TRUE)
par(mfrow=c(1, 2))
hist(mean.diff, breaks = "Scott", col = "lightblue", main = "Bootstrap Distribution", xlab = substitute(paste(bar(x)[1],"*", - bar(x)[2],"*")), freq= FALSE)
abline(v = mean(AK.weight) - mean(WY.weight), col = "red")
curve(dnorm(x, mean(mean.diff), sd(mean.diff)), col ="blue", add= TRUE, lwd = 2)
qqnorm(mean.diff, col ="blue")
qqline(mean.diff, col = "red")
par(opar)
BIAS <- mean(mean.diff) - (mean(AK.weight) - mean(WY.weight))
BIAS
CI <- quantile(mean.diff, prob = c(.025, .975))
CI
```

```{r GirlsGGplot}
ggplot(data = data.frame(x = mean.diff), aes(x = x)) + geom_density(fill = "skyblue") + labs(x = substitute(paste(bar(x)[1],"*", - bar(x)[2],"*"))) + theme_bw()
```

> ANSWER: The mean of the bootstrap distribution of the difference of means ($\hat{\mu}_{\bar{x}^*_1 - \bar{x}^*_2}$) is `r mean(mean.diff)` while the standard error of the bootstrap distribution of means ($\hat{\sigma}_{\bar{x}^*_1 - \bar{x}^*_2}$) is `r sd(mean.diff)`.  We are 95% confident that the mean weight of baby girls born in Alaska are `r CI[1]` to `r CI[2]` grams greater than baby girls born in Wyoming.
  
  * What is the bootstrap estimate of the bias?  What fraction of the bootstrap standard error does it represent?
  
> ANSWER: The bootstrap estimate of the bias $(Bias_{boot}[\bar{x}^*_1 - \bar{x}^*_2] = E[\bar{x}^*_1 - \bar{x}^*_2] - [\bar{x}_1 - \bar{x}_2])$ is `r BIAS`.  The $Bias_{boot}[\bar{x}^*_1 - \bar{x}^*_2]$ represents `r abs(BIAS/sd(mean.diff)*100)`% of the standard error of the bootstrap.
  
  *  Conduct a permutation test to calculate the difference in mean weights and state your conclusion.
  
```{r GirlsPer}
ANS <- tapply(Girls$Weight, Girls$State, mean)
ANS
obs.mean.diff <- ANS[1] - ANS[2]
names(obs.mean.diff) <- NULL
obs.mean.diff
R <- 50000
Weight <- subset(Girls, select = Weight, drop = TRUE)
result <- numeric(R)  # space to save the random differences
for (i in 1:R){
  # sample of size 40, from 1 to 80, without replacement
  index <- sample(80, size = 40, replace = FALSE)
  result[i] <- mean(Weight[index]) - mean(Weight[-index])
}
hist(result, col = "blue", main = "", xlab = expression(bar(x)[1]- bar(x)[2]))
abline(v = obs.mean.diff, col = "red")
pvalue <- (sum(abs(result) >= abs(obs.mean.diff) + 1))/(R + 1) # p-value
pvalue  # results will vary
```
The _p_-value for testing $H_0: \mu_{AK} = \mu_{WY}$ versus $H_A: \mu_{AK} \neq \mu_{WY}$ is `r pvalue`.  The small _p_-value suggests that the mean weight of babies born in Alaska is not the same as the mean weight of babies born in Wyoming.
  
  *  For what population(s) if any, does this conclusion hold?  Explain.
  
> ANSWER:  The results should hold for single birth babies born in both Alaska and Wyoming with a gestation period of at least 37 weeks.  
