---
title: "Chapter Five Homework"
author: "Your Name Here"
date: '`r format(Sys.time(), "%A, %b %d, %Y - %X")`'
output: bookdown::html_document2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = NA, fig.align = "center", warning = FALSE, message = FALSE)
library(tidyverse)
```

1. Consider a population that has a normal distribution with mean $\mu = 36$, standard deviation $\sigma = 8$

    (a) The sampling distribution of $\bar{X}$ for samples of size 200 will have what distribution, mean, and standard error?
    
    (b) Use `R` to draw a random sample of size $200$ from this population. Conduct EDA on your sample.
    
    (c) Compute the bootstrap distribution for your sample mean, and note the bootstrap mean and standard error.
    
    (d) Compare the bootstrap distribution to the theoretical sampling distribution by creating a table like Table 5.2.
    
    (e) Repeat parts a-d for sample sizes of $n = 50$ and $n = 10$. Carefully describe your observations about the effects of sample size on the bootstrap distribution.
 
------------ 

Your answers:

a.  

b.

```{r}
# Your code here
set.seed(13)

```



c.

```{r}
# Your code here

```

d.

```{r}
# Your code here

```

----------------

e.

Parts a-d for $n = 10$:

a.  

b.

```{r}
# Your code here

```

c.

```{r}
# Your code here
set.seed(31)

```

d.

```{r}
# Your code here

```

------------

Parts a-d for $n = 50$:

a.  

b.

```{r}
# Your code here

```


c.

```{r}
# Your code here
set.seed(31)

```

d.

```{r}
# Your code here

```

 
**Your Answer here.**

------------ 

2. We investigate the bootstrap distribution of the median. Create random samples of size $n$ for various $n$ and bootstrap the median. Describe the bootstrap distribution.
Change the sample sizes to 36 and 37; 200 and 201; 10,000 and 10,001. Note the similarities/dissimilarities, trends and so on. Why does the parity of the sample size matter? 

```{r, label = "Label1415", fig.cap = "Histograms of bootstrapped median values"}
set.seed(31)
ne <- 14 # n even
no <- 15 # n odd

wwe <- rnorm(ne) # draw random sample of size ne
wwo <- rnorm(no) # draw random sample of size no

N <- 10^4
even.boot <- numeric(N) # save space
odd.boot <- numeric(N)
for (i in 1:N)
{
  x.even <- sample(wwe, ne, replace = TRUE)
  x.odd <- sample(wwo, no, replace = TRUE)
  even.boot[i] <- median(x.even)
  odd.boot[i] <- median(x.odd)
}

Median <- c(even.boot, odd.boot)
Parity <- rep(c("n = 14", "n = 15"), each = N)
DF <- data.frame(Median = Median, Parity = Parity)

ggplot(data = DF, aes(x = Median)) + 
  geom_histogram(fill = "lightblue", color = "black") + 
  theme_bw() + 
  facet_grid(Parity ~.)
```

--------------

```{r, label = "Label3637", fig.cap = "Histograms of bootstrapped median values", cache = TRUE}
set.seed(31)
# Your code here

```


```{r, label = "Label200201", fig.cap = "Histograms of bootstrapped median values", cache = TRUE}
set.seed(31)
# Your code here

```


```{r, label = "Label10000", fig.cap = "Histograms of bootstrapped median values", cache = TRUE}
set.seed(31)
# Your code here

```


------------ 

Your answer:

 
------------      

3. Import the data from data set `Bangladesh`. In addition to arsenic concentrations for 271 wells, the data set contains cobalt and chlorine concentrations.
    
    (a) Conduct EDA on the chlorine concentrations and describe the salient features.
    
    (b) Bootstrap the mean.
    
    (c) Find and interpret the 95% bootstrap percentile confidence interval.
    
    (d) What is the bootstrap estimate of the bias? What fraction of the bootstrap standard error does it represent?
    
```{r, label = "Bang"}
Bangladesh <- read.csv("http://www1.appstate.edu/~arnholta/Data/Bangladesh.csv")
head(Bangladesh)
```

The `Chlorine` variable has some missing values. The following code will remove these entries:
    
```{r}
chlorine <- subset(Bangladesh, select = Chlorine, subset = !is.na(Chlorine), drop = TRUE)
```

------------ 

Your answers:

a.

```{r}
# Your code here

```



b.

```{r}
# Your code here

```


c.

```{r}
# Your code here

```



d.


 
------------  

4. Consider Bangladesh chlorine (concentration). Bootstrap the trimmed mean (say, trim the upper and lower 25%) and compare your results with the usual mean (previous exercise).

------------ 

Your answer:

```{r}
# Your code here

```


 
------------  

5. The data set `FishMercury` contains mercury levels (parts per million) for 30 fish caught in lakes in Minnesota.

    (a) Create a histogram or boxplot of the data. What do you observe?
    
    (b) Bootstrap the mean and record the bootstrap standard error and the 95% bootstrap percentile interval.
    
    (c) Remove the outlier and bootstrap the mean of the remaining data. Record the bootstrap standard error and the 95% bootstrap percentile interval.
    
    (d) What effect did removing the outlier have on the bootstrap distribution, in particular, the standard error?
    
```{r, label = "Fish"}
FishMercury <- read.csv("http://www1.appstate.edu/~arnholta/Data/FishMercury.csv")
head(FishMercury)
```    

------------ 

Your answers:

a.

```{r}
# Your code here
```

Note that there is one value (`r max(FishMercury$Mercury)`) very far removed from the rest of the values.

b.

```{r}
# Your code here
```



c.

```{r}
# Your code here
```


d.
 


------------  
    
6. In section 3.3, we performed a permutation test to determine if men and women consumed, on average, different amounts of hot wings.

    (a) Bootstrap the difference in means and describe the bootstrap distribution.
    
    (b) Find a 95% bootstrap percentile confidence interval for the difference of means and give a sentence interpreting this interval.
    
    (c) How do the bootstrap and permutation distribution differ?
    
------------ 

```{r}
BeerWings <- read.csv("http://www1.appstate.edu/~arnholta/Data/Beerwings.csv")
head(BeerWings)
```

Your answers:

a.

```{r}
# Your code here

```




b.

```{r}
# Your code here
```



c.


 
------------      
    
7. Import the data from `Girls2004` (see Section 1.2).

    (a) Perform some exploratory data analysis and obtain summary statistics on the weight of baby girls born in Wyoming and Arkansas (do seperate analyses for each state).
    
    (b) Bootstrap  the difference in means, plot the distribution, and give the summary statistics. Obtain a 95% bootstrap percentile confidence interval and interpret this interval.
    
    (c) What is the bootstrap estimate of the bias? What fraction of the bootstrap standard error does it represent?
    
    (d) Conduct a permutation test to calculate the difference in mean weights and state your conclusion?
    
    (e) For what population(s), if any does this calculation hold? Explain?

```{r, label = "Girls"}
Girls2004 <- read.csv("http://www1.appstate.edu/~arnholta/Data/Girls2004.csv")
head(Girls2004)
```

------------ 

Your answers:

a.

```{r}
# Your code here
# Part a.

```


b.

```{r}
# Your code here
```


c.

```{r}
# Your code here

```



d.

```{r}
# Your code here

```



e.


 
------------  
    
8. Do chocolate and vanilla ice creams have the same number of calories? The data set `IceCream` contains calorie information for a sample of brands of chocolate and vanilla ice cream. Use the bootstrap to determine whether or not  there is a difference in the mean number of calories.

```{r, label = "IceCream"}
IceCream <- read.csv("http://www1.appstate.edu/~arnholta/Data/IceCream.csv")
head(IceCream)
```

------------ 

Your answer:

```{r}
# Your code here

```



------------  

9. Import the data from Flight Delays Case Study in Section 1.1 data into `R`. Although the data are on all UA and AA flights flown in May and June of 2009, we will assume these represent a sample from a larger population of UA and AA flights flown under similiar circumstances. We will consider the ratio of the means of the flight delay lengths, $\mu_{\text{UA}} / \mu_{\text{AA}}$.

    (a) Perform some exploratory data analysis on flight delay lengths for each of UA and AA flights.
    
    (b) Bootstrap the mean of flight delay lengths for each airline seperately and describe the distribution.
    
    (c) Bootstrap the ratio of means. Provide plots of the bootstrap distribution and describe the distribution.
    
    (d) Find the 95% bootstrap percentile interval for the ratio of means. Interpret this interval.
    
    (e) What is the bootstrap estimate of the bias? What fraction of the bootstrap standard error does it represent?
    
    (f) For inference in this text, we assume that the observations are independent. Is that condition met here? Explain.
    
```{r, label = "FD"}
FlightDelays <- read.csv("http://www1.appstate.edu/~arnholta/Data/FlightDelays.csv")
```

------------ 

Your answers:

a.

```{r}
# Your code here

```


b.

```{r}
# Your code here

```


c.

```{r}
# Your code here

```



d.

```{r}
# Your code here

```



e.

```{r}
# Your code here

```



f.

 
------------  
    
10. Two college students collected data on the price of hardcover textbooks from two disciplinary areas: Mathematics and the Natural Sciences, and the Social Sciences (Hien and Baker (2010)). The data are in the file `BookPrices`.

    (a) Perform some exploratory data analysis on book prices for each of the two disciplinary areas.
    
    (b) Bootstrap the mean of the book price for each area separately and describe the distributions.
    
    (c) Bootstrap the ratio of means. Provide plots of the bootstrap distribution and comment.
    
    (d) Find the 95% bootstrap percentile interval for the ratio of means. Interpret this interval.
    
    (e) What is the bootstrap estimate of the bias? What fraction of the bootstrap standard error does it represent?
    
```{r, label = "BookPrices"}
BookPrices <- read.csv("http://www1.appstate.edu/~arnholta/Data/BookPrices.csv")
```

------------ 

Your answers:

a.

```{r}
# Your code here

```


b.

```{r}
# Your code here

```



c.

```{r}
# Your code here

```



d.

```{r}
# Your code here

```


e.

```{r}
# Your code here

```

 
------------  