---
title: "Sampling Distributions"
author: "Alan Arnholt"
date: "11/7/2017"
output: bookdown::html_document2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = NA, warning = FALSE, message = FALSE)
```

# Uniform Distribution

## Definition

A random variable $X$ has a uniform distribution on the interval $[a, b]$ (for $a < b$) if its **pdf** is

\begin{equation}
f(x) = 
\begin{cases}
\frac{1}{b-a},  & a \leq x \leq b \\
0,  & \text{otherwise.}
\end{cases}
(\#eq:unif)
\end{equation}

If $X \sim \text{Unif}(a, b)$, then 

\begin{equation}
E(X) = \frac{a + b}{2}
(\#eq:EU)
\end{equation}

\begin{equation}
V(X) = \frac{(b-a)^2}{12}
(\#eq:VU)
\end{equation}

## Example

Suppose $X \sim \text{Unif}(10, 20)$. 

a.  Find the $E(X)$ and $V(X)$ using the definitions of $E(X)$ and $V(X)$.

```{r}
# part a.
xfx <- function(x){x/10}
EX <- integrate(xfx, 10, 20)$value
EX
x2fx  <- function(x){(x - EX)^2/10}
VX <- integrate(x2fx, 10, 20)$value
VX
```
      
b. Find the $E(X)$ and $V(X)$ using the short cut formulas in \@ref(eq:EU) and \@ref(eq:VU).

$E(X) = \frac{a+b}{2}=\frac{10 + 20}{2} = 15$, and $V(X) = \frac{(b-a)^2}{12} = \frac{(20 - 10)^2}{12} = \frac{100}{12} = `r 100/12`.$
      
c. Simulate 10,000 values of the random variable and estimate $E(X)$ and $V(X)$.

```{r}
# part c.
set.seed(89)
sims <- 10^4
X <- runif(sims, 10, 20)
EX <- mean(X)
VX <- var(X)
c(EX, VX)
```
d. Find $E(\bar{X})$ and $V(\bar{X})$ for $n = 8$ exactly and via simulation.

$E(\bar{X}) = \mu_{\bar{X}} = \mu_X = 15$, $V(\bar{X}) = \frac{\sigma^2_{X}}{n} = \frac{\frac{100}{12}}{8} = `r 100/12/8`$ 

```{r}
set.seed(46)
sims <- 10^4
n <- 8
a <- 10
b <- 20
xbar <- numeric(sims)
for(i in 1:sims){
  X <- runif(n, a, b)
  xbar[i] <- mean(X)
}
mean(xbar)
var(xbar)
```

# Exponential

## Problem

Let $X_1, X_2, \ldots, X_{20} \overset{i.i.d}\sim\text{Exp}(\lambda = 2)$.  Let $Y = \sum_{i=1}^{20}X_i.$

a.  Simulate the sampling distribution of $Y$ in `R`.

```{r}
sims <- 10^4
Y <- numeric(sims)
for(i in 1:sims){
 Y[i] <- sum(rexp(20, 2))
}
EY <- mean(Y)
VY <- var(Y)
c(EY, VY)
mean(Y <= 10)
```

