---
title: "Chapter 4 Problem 22"
author: "Alan Arnholt"
date: "March 27, 2014"
output:
  beamer_presentation:
    keep_tex: yes
    theme: Madrid
incremental: yes
---
## Problem 22

Let $X_1, X_2, \ldots X_n \overset{iid}\sim F$ with corresponding pdf $f(x) = 3x^2, 0\leq x \leq 1.$

\begin{enumerate}[(a)]
\item Find the pdf for $X_{\textrm{min}}.$ \label{parta}
\item Find the pdf for $X_{\textrm{max}}.$ \label{partb}
\item If $n = 10,$ find the probability that the largest value, $X_{\textrm{max}},$ is greater than 0.92 \label{partc}
\item if $n = 10,$ find the expected value of $X_{\textrm{max}}.$ \label{partd}
\end{enumerate}

## $f_{\textrm{min}}(x)$

\eqref{parta}  The pdf for $X_{\textrm{min}}$ is:
\begin{equation}\label{Xmin}
f_{\textrm{min}}(x) = n\left(1 - F(X)\right)^{n-1}f(x)
\end{equation}

Since $F(x) = \int_{0}^{x}3t^2\,dt = x^3, 0\leq x \leq 1,$ it follows that 
$$f_{\textrm{min}}(x) = n\left(1 - x^3\right)^{n-1}3x^2,  0\leq x \leq 1$$

## $f_{\textrm{max}}(x)$

\eqref{partb} The pdf for $X_{\textrm{max}}$ is:
\begin{equation}\label{Xmax}
f_{\textrm{max}}(x) = nF^{n-1}(x)f(x)
\end{equation}

Since $F(x) = \int_{0}^{x}3t^2\,dt = x^3, 0\leq x \leq 1,$ it follows that 
$$f_{\textrm{max}}(x) = n\left(x^3\right)^{n-1}3x^2,  0\leq x \leq 1$$

## $P(X_{\textrm{max}} > 0.92)$

\eqref{partc} $$f_{\textrm{max}}(x) = 10\left(x^3\right)^{10-1}3x^2 = 30x^{29},  0\leq x \leq 1$$

```{r label = FUNC, comment = NA}
f <- function(x){30*x^29}
ans <- integrate(f, 0.92, 1)$value
ans
```

$$P(X_{\textrm{max}} > 0.92) = \int_{0.92}^{1}30x^{29}\,dx = `r ans`.$$

## $E[X_{\textrm{max}}]$

$$E[X_{\textrm{max}}] = \int_{0}^{1}x_{\textrm{max}} f_{\textrm{max}}(x) = \int_{0}^{1}30x^{30}$$

```{r comment = NA}
f1 <- function(x){30*x^30}
ans <- integrate(f1, 0, 1)$value
ans
```

$$E[X_{\textrm{max}}] = `r ans`$$

## What does the pdf look like?

```{r comment = NA, fig.height = 3, fig.width = 5}
curve(f, 0, 1, n = 500, col ="purple")
abline(v = ans, lty = "dashed", col = "red")
```