---
title: "Least Squares Regression"
author: "Alan T. Arnholt"
date: last-modified
date-format: "[Last modified on] MMMM DD, YYYY HH:mm:ss zzz"
format: 
  html: default
  pdf: default
editor: source
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = NA, warning = FALSE, message = FALSE, fig.align = "center")
```

# Correlation

The **correlation coefficient**, denoted by $r$, measures the direction and strength of the linear relationship between two numerical variables.  Is is given by the equation

$$
r = \frac{1}{(n-1)}\sum_{i=1}^n\left(\frac{x_i - \bar{x}}{s_x}\right)\left(\frac{y_i - \bar{y}}{s_y}\right)
$${#eq-cor}

Following are the high school GPAs and the college GPAs at the end of the freshman year for ten different students from the `Gpa` data set of the `BSDA` package.

```{r, echo = FALSE}
library(BSDA)
knitr::kable(Gpa)
```

Create a scatterplot and then comment on the relationship between the two variables.

::: {.callout-caution icon="false" title="R Code"}
```{r}
#| label: "fig-sp1"
#| fig-cap: "Scatterplot of College GPA versus High School GPA"
library(tidyverse)
library(BSDA)
ggplot(data = Gpa, aes(x = hsgpa, y = collgpa)) + 
  labs(x = "High School GPA", y = "College GPA") +
  geom_point() + 
  theme_bw()
```
:::

The college GPA is the response variable and is labeled on the vertical axis.  The scatterplot in @fig-sp1 shows that the college GPA increases as the high school GPA increases.  In fact, the the dots appear to cluster along a straight line.  The correlation coefficient is $r = `r round(cor(Gpa$hsgpa, Gpa$collgpa), 3)`$, which indicates that a straight line is a reasonable relationship between the two variables.

* Compute the correlation coefficient using @eq-cor.

::: {.callout-caution icon="false" title="R Code"}
```{r}
head(Gpa)
values <- Gpa |> 
  mutate(y_ybar = collgpa - mean(collgpa), 
         x_xbar = hsgpa - mean(hsgpa), 
         zx = x_xbar/sd(hsgpa), 
         zy = y_ybar/sd(collgpa),
         zxzy = zx*zy)
knitr::kable(values)
#
values |>  
  summarize(r = (1/9)*sum(zxzy))
```

Using the build in `cor()` function:

```{r}
Gpa |>  
  summarize(r = cor(collgpa, hsgpa))
```
:::

Note:

```{r}
p1 <- ggplot(data = Gpa, aes(x = hsgpa, y = collgpa)) + 
  geom_point() + 
  theme_bw()
p2 <- ggplot(data = values, aes(x = zx, y = zy)) + 
  geom_point() + 
  theme_bw()
library(gridExtra)
grid.arrange(p1, p2, ncol = 1, nrow = 2)
# Or better yet
library(patchwork)
p1 / p2
p1 + p2
p1 + gt::gt(Gpa)
```


# Least Squares Regression

The equation of a straight line is

$$y = b_0 + b_1x$${#eq-sl}

where $b_0$ is the $y$-intercept and $b_1$ is the slope of the line.  From the equation of the line that best fits the data,

$$\hat{y} = b_0 + b_1x$${#eq-lr}

we can compute a predicted $y$ for each value of $x$ and then measure the error of the prediction.  The error of the prediction, $e_i$ (also called the residual) is the difference in the actual $y_i$ and the predicted $\hat{y}_i$.  That is, the residual associated with the data point $(x_i, y_i)$ is

$$e_i = y_i - \hat{y}_i.$${#eq-res}

The least squares regression line is

$$\hat{y} = b_0 + b_1x$${#eq-lsr}

where

$$
b_1 = \frac{\sum(x_i - \bar{x})(y_i - \bar{y})}{\sum(x_i - \bar{x})^2} = r\frac{s_y}{s_x}
$${#eq-b1}

and 

$$
b_0 = \bar{y} - b_1\bar{x}
$${#eq-b0}

Find the least squares regression line $\hat{y} = b_0 + b_1x$ for the `Gpa` data.

::: {.callout-caution icon="false" title="R Code"}
```{r}
Gpa |>  
  summarize(b1 = cor(hsgpa, collgpa)*sd(collgpa)/sd(hsgpa),
            b0 = mean(collgpa) - b1*mean(hsgpa)) -> ans1
ans1 |> 
  knitr::kable()
```
$$
\widehat{\text{collgpa}} = `r ans1$b0` + `r ans1$b1` \times \text{hsgpa}
$${#eq-lsl}

:::

The coefficients are also computed when using the `lm()` function.

::: {.callout-caution icon="false" title="R Code"}
```{r}
mod1 <- lm(collgpa ~ hsgpa, data = Gpa)
mod1
summary(mod1)
```

$$
\widehat{\text{collgpa}} = `r mod1$coef[1]` + `r mod1$coef[2]` \times \text{hsgpa}
$${#eq-lsl2}

Or

$$
\widehat{\text{collgpa}} = `r summary(mod1)$coef[1, 1]` + `r summary(mod1)$coef[2, 1]` \times \text{hsgpa}
$${#eq-lsl3}

```{r}
knitr::kable(summary(mod1)$coef)
library(moderndive)
get_regression_table(mod1)
# Spruce up some
get_regression_table(mod1) |> 
  knitr::kable()
```

$$
\widehat{\text{collgpa}} = `r get_regression_table(mod1)[1, 2]` + `r get_regression_table(mod1)[2, 2]` \times \text{hsgpa}
$${#eq-lsl4}

```{r}
broom::tidy(mod1, conf.int = TRUE) |>
  knitr::kable()
# GOF measures
broom::glance(mod1) |>
  knitr::kable()
```

$$
\widehat{\text{collgpa}} = `r broom::tidy(mod1)[1, 2]` + `r broom::tidy(mod1)[2, 2]` \times \text{hsgpa}
$${#eq-lsl5}

:::

Find the residuals for `mod1`.

::: {.callout-caution icon="false" title="R Code"}
```{r}
get_regression_points(mod1)
# Spruced up
get_regression_points(mod1) |> 
  knitr::kable()
# or broom
broom::augment(mod1)
# Spruced up
broom::augment(mod1) |> 
  knitr::kable()
```
:::

Add the least squares line to the scatterplot for `collgpa` versus `hsgpa`.

::: {.callout-caution icon="false" title="R Code"}
```{r}
ggplot(data = Gpa, aes(x = hsgpa, y = collgpa)) + 
  geom_point() + 
  geom_smooth(method = "lm", se = FALSE) + 
  labs(x = "High School GPA", y = "College GPA") +
  theme_bw() 
```
:::

## Using the model to predict

What does the least squares regression model (`mod1`) predict for `collgpa` for a student with a 3.3 `hsgpa`?

```{r}
-0.950366 + 1.346999*3.3
mod1$coef[1] + mod1$coef[2]*3.3
predict(mod1, newdata = tibble(hsgpa = 3.3))
get_regression_points(mod1, newdata = tibble(hsgpa = 3.3), digits = 7) -> T3
T3
T3$collgpa_hat
T3[1, 3]
T3[1, 3] |> pull()
```


## Assessing the fit

```{r}
# R^2 plot
broom::augment(mod1) |> 
  ggplot(aes(x = collgpa, y = .fitted)) + 
  geom_point() +
  geom_abline(color = "red") + 
  tune::coord_obs_pred() +
  theme_bw()
```


::: {.callout-caution icon="false" title="R Code"}
```{r, fig.width = 8, fig.height = 4}
library(ggfortify)
autoplot(mod1, ncol = 2, nrow = 1, which = 1:2) + theme_bw()
```
:::

