---
title: "Three S's"
author: "Alan T. Arnholt"
date: '`r format(Sys.time(), "%A, %B %d, %Y - %X.")`'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = NA, fig.align = "center")
```

## Using R Markdown

Using R Markdown allows us to write both text and code in the same document.  Use R code chunks to insert code:

````markdown
`r ''````{r}
# Some code
x <- 1:10
```
````

Use inline `R` to write answers inline using the following format: `` ``r ''`r R_CODE` ``.  For example, to compute the mean of the values 1, 3, 5, and 7, one might can use `` ``r ''`r mean(c(1, 3, 5, 7))` ``.  The mean of 1, 3, 5, and 7 is `r mean(c(1, 3, 5, 7))`.

### Using packages

To use functions in packages such as `psych`, one must either specify the package by prepending the function with the package name and two colons or load the package using the command `library(PackageName)`.

Consider using the function `describe` on the `mtcars` data set.

```{r, error = TRUE}
describe(mtcars) # psych has not been loaded!
psych::describe(mtcars)
```

### Characterizing `qsec`

* Shape

```{r}
hist(mtcars$qsec, col = "blue", freq = FALSE, 
      main = "Histogram of time to travel quarter mile",
      xlab = "time in seconds")
```

* Center

```{r}
Mean <- mean(mtcars$qsec)
Mean
```

* Spread

```{r}
SD <- sd(mtcars$qsec)
SD
```

The distribution of `qsec` is unimodal and symmetric with a mean of `r round(Mean,2)` seconds and a standard deviation of `r round(SD, 2)` seconds.

## Superimposing a Normal Distribution

```{r}
hist(mtcars$qsec, col = "blue", freq = FALSE, 
      main = "Histogram of time to travel quarter mile",
      xlab = "time in seconds", xlim = c(13, 23))
curve(dnorm(x, Mean, SD), 13, 23, col = "purple", add = TRUE, lwd = 3)
```

## Using `ggplot2`

```{r}
library(ggplot2)
ggplot(data = mtcars, aes(x = qsec, ..density..)) +
  geom_histogram(binwidth = 1, fill = "blue", color = "black") + 
  xlim(Mean - 3.5*SD, Mean + 3.5*SD) +
  labs(x = "Time in seconds") +
  geom_density(fill = "red", alpha = 0.4) +
  stat_function(fun = dnorm, args = list(mean = Mean, sd = SD), 
                inherit.aes = FALSE, size = 2, color = "purple") +
  theme_bw()
```

## Tests of Significance

1. Hypotheses --- State the null and alternative hypotheses.

2. Test Statistic

3. Rejection Region Calculations

4. Statistical Conclusion

5. English Conclusion

### Example 9.12 from PASWR2

A bottled water company acquires its water from two independent sources, X and Y.  The company suspects that the sodium content in the water from source X is less than the sodium content from source Y.  An independent agency measures the sodium content in 20 samples from source X and 10 samples from source Y and stores them in data frame `WATER` of the `PASWR2` package.  Is there statistical evidence to suggest the average sodium content in the water from source X is less than the average sodium content in Y?

**Solution:** To solve this problem, start by verifying the reasonableness of the normality assumption.

```{r, warning = FALSE, message = FALSE}
library(PASWR2)     # load the PASWR2 package
library(ggplot2)    # load the ggplot2 package
library(lsr)        # load the lsr package
library(DescTools)  # load the DescTools package
boxplot(sodium ~ source, data = WATER)
ggplot(data = WATER, aes(x = source, y = sodium)) + 
geom_boxplot() +
theme_bw()
LeveneTest(sodium ~ source, data = WATER)
ggplot(data = WATER, aes(sample = sodium, color = source)) + 
stat_qq() + 
theme_bw()
CohenD(WATER$x, WATER$y, na.rm = TRUE)
cohensD(formula = sodium ~ source, data = WATER)
```

1. **Hypotheses** --- $H_0: \mu_X - \mu_Y = 0$ versus $H_1: \mu_X - \mu_Y <0$

2. **Test Statistic** ---The test statistic is $\bar{X} - \bar{Y}$.  The standardized test statistic under the assumptioon that $H_0$ is true and its approximate distribution are

$$\frac{\left[(\bar{X}-\bar{Y} - \delta_0) \right]}{\sqrt{\frac{S_X^2}{n_x}+\frac{S_Y^2}{n_Y}}} \overset{\bullet}{\sim} t_{\nu}$$

3. **Rejection Region Calculations** --- $P(t_{obs} < t_{0.05, 22.069}) = `r qt(0.05, 22.069)`.$

```{r}
TR <- t.test(sodium ~ source, data = WATER, alternative = "less")
TR
```

4. **Statistical Conclusion** --- Since the p-value is `r TR$p.value`, reject the null hypothesis.

5. **English Conclusion** --- There is evidence to suggest the average sodium content for source X is less than the average sodium content for source Y.