---
title: "Reading In Data"
author: "Alan T. Arnholt"
date: 'Last compiled: `r format(Sys.time(), "%B %d, %Y at %X")`'
output: bookdown::html_document2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, comment = NA, fig.align = "center")
library(tidyverse)
library(readr)
library(data.table)
```

# Data

This section presents a data set that shows how different data types should be read into `R` as well as several functions that are useful for working with different types of `R` objects.  Consider the data stored as a CSV file at

https://raw.githubusercontent.com/alanarnholt/Data/master/POPLAR3.CSV


The following description of the data is from `Minitab 15`:

```
In an effort to maximize yield, researchers designed an experiment to determine how two factors, Site and Treatment, influence the Weight of four-year-old poplar clones. They planted trees on two sites: Site 1 is a moist site with rich soil, and Site 2 is a dry, sandy site. They applied four different treatments to the trees: Treatment 1 was the control (no treatment);   Treatment 2 used fertilizer; Treatment 3 used irrigation; and Treatment 4 use both fertilizer and irrigation. To account for a variety of weather conditions, the researchers replicated the data by planting half the trees in Year 1, and the other half in Year 2.
```

## Base R

The data from `Poplar3.CSV` is read into the data frame `poplar` using the `read.csv()` function, and the first five rows of the data frame are shown using the function `head()` with the argument `n = 5` to show the first five rows of the data frame instead of the default `n = 6` rows.


```{r}
site <- "https://raw.githubusercontent.com/alanarnholt/Data/master/POPLAR3.CSV"
poplar <- read.csv(file = url(site))
head(poplar, n = 5)  # show first five rows
```

When dealing with imported data sets, it is always good to examine their contents using functions such as `str()` and `summary()`, which show the structure and provide appropriate summaries, respectively, for different types of objects.

```{r}
str(poplar)
summary(poplar)
```

From typing `str(poplar)` at the `R` prompt, one can see that all seven variables are either integer or numeric.  From the description, the variables `Site` and `Treatment` are factors.  Further investigation into the experiment reveals that `year` and `Age` are factors as well. Recall that factors are an extension of vectors designed for storing categorical information.   The results of `summary(poplar)` indicate the minimum values for `Diameter`, `Height`, and `Weight` are all `-99`, which does not make sense unless one is told that a value of `-99` for these variables represents a missing value.  Once one understands that the variables `Site`, `Year`, `Treatment`, and `Age` are factors and that the value `-99` has been used to represent missing values for the variables `Diameter`, `Height`, and `Weight`, appropriate arguments to `read.csv()` can be entered.  The data is now read into the object `poplarC` using `na.strings = "-99"` to store the `NA` values correctly.  The argument `colClasses=` requires a vector that indicates the desired class of each column.

```{r}
poplarC <- read.csv(file = url(site), na.strings = "-99", 
    colClasses = c(rep("factor", 3), rep("numeric", 3), "factor"))
str(poplarC)
```

In the event different values (999, 99, 9999) for different variables (`var1`, `var2`, `var3`) are used to represent missing values in a data set, the argument `na.strings=` will no longer be able to solve the problem directly.  Although one can pass a vector of the form `na.strings = c(999, 99, 9999)`, this will simply replace all values that are 999,  99, or 9999 with `NA`s.  If the first variable has a legitimate value of 99, then it too would be replaced with an `NA` value.  One solution for this problem in general is to read the data set into a data frame (`DF`), to assign the data frame to a different name so that the cleaned up data set is not confused with the original data, and to use filtering to assign `NA`s to values of `var1`, `var2`, and `var3` that have entries of 999, 99, and 999, respectively.

```{r eval = FALSE}
DF <- read.table(file=url(site), header=TRUE)
df <- DF
df[df$var1==999,  "var1"] = NA
df[df$var2==99,   "var2"] = NA
df[df$var3==9999, "var3"] = NA
```

Once a variable has its class changed from `int` to `factor`, labeling the levels of the factor can be accomplished without difficulties.  To facilitate analysis of the `poplarC` data, labels for the levels of the variables `Site` and `Treatment` are assigned.

```{r c01LevelsSite}
poplarC$site <- factor(poplarC$site, labels = c("Moist", "Dry"))
TreatmentLevels <- c("Control", "Fertilizer", "Irrigation", "FertIrriga")
poplarC$Treatment <- factor(poplarC$treatment, labels = TreatmentLevels)
str(poplarC)
```

## Using `readr` functions

```{r}
poplarR <- read_csv(file = url(site))
head(poplarR, n = 5)  # show first five rows
summary(poplarR)
#
poplarR1 <- read_csv(file = url(site), na = "-99")
summary(poplarR1)
str(poplarR1)
```

## Using `data.table` function `fread()`
 
```{r}
poplarF <- fread(input = site, na.strings = "-99")
str(poplarF)
summary(poplarF)
```
 
# Graphing Now

Basic scatterplot:

```{r}
ggplot(data = poplarC, mapping = aes(x = diameter, y = height)) + 
  geom_point() 
```

```{r}
ggplot(data = poplarC, mapping = aes(x = diameter, y = height, color = treatment)) + 
  geom_point() +
  theme_bw() + 
  geom_smooth(se = FALSE)
```

```{r}
ggplot(data = poplarC, mapping = aes(x = diameter, y = height, color = treatment)) + 
  geom_point() +
  theme_bw() + 
  geom_smooth(se = FALSE) + 
  facet_grid(cols = vars(age), rows = vars(site))
```


