---
title: "5NG - Five Named Graphs"
author: "Alan Arnholt"
date: last-modified
date-format: "[Last modified on] MMMM DD, YYYY HH:mm:ss zzz"
format: 
  html:
    code-fold: true
    theme:
      light: litera
      dark: cyborg
---

# Scatterplots and overplotting

Consider the data set `evals` from the `moderndive` package.  First load both `tidyverse` and `moderndive`.

```{r}
#| warning: false
#| message: false
library(tidyverse)
library(moderndive)
```

What is `evals` measuring?

```{r}
?evals
```

The data are gathered from end of semester student evaluations for a sample of 463 courses taught by 94 professors from the University of Texas at Austin. In addition, six students rate the professors' physical appearance. The result is a data frame where each row contains a different course and each column has information on either the course or the professor.

## Create a scatterplot of `score` versus `bty_avg`

```{r}
#| code-fold: true
#| code-summary: "I can't wait to see the code"
ggplot(data = evals, aes(x = bty_avg, y = score)) + 
  labs(x = "Teacher's average beauty score", 
       y = "Teacher's average course evaluation") -> p1
p1 +geom_point() 
```

Let's look at a few ways to fix some of the over plotting.  

1. Use a different geom (`geom_jitter`).
2. Use `alpha`
3. Use `geom_count()` or `stat_sum()`

```{r}
#| code-fold: true
#| code-summary: "Show the code"
p1 + geom_jitter()
p1 + geom_point(alpha = .25)
p1 + geom_count()
p1 + stat_sum()
p1 + geom_count(alpha = .25)
p1 + geom_count(alpha = 0.25) + scale_size(range = c(1, 10))
p1 + stat_sum(alpha = 0.25, aes(size = after_stat(prop)))
```

```{r}
#| warning: false
p1 + 
  geom_count(alpha = 0.25, aes(color = gender)) + 
  scale_size(range = c(1, 10)) +
  geom_smooth(aes(color = gender), method = "loess", se = FALSE) + 
  labs(color = "Sex") + 
  scale_color_manual(values = c("female" = "pink", "male" = "darkblue")) + 
  theme_classic()
```


Consider the `early_january_weather` from the `moderndive` package.

```{r}
#| code-fold: false
#| echo: false
library(moderndive)
knitr::kable(head(early_january_weather))
```

## Linegraphs via `geom_line()`

```{r}
ggplot(data = early_january_weather, aes(x = time_hour, y = temp)) + 
  geom_point(size = 0.2, color = "blue") + 
  geom_line(color = "pink") + 
  labs(x = "Date and time",
       y = "Temperature (in Fahrenheit)")

```

## Histograms via `geom_histogram()`

```{r}
#| warning: false
library(palmerpenguins)
penguins |> 
  ggplot(aes(x = bill_length_mm)) + 
  geom_histogram() + 
  labs(x = "Penguin bill length (in mm)")
```

```{r}
#| warning: false
penguins |> 
  ggplot(aes(x = bill_length_mm)) + 
  geom_histogram(binwidth = 3, color = "black", fill = "lightblue") + 
  labs(x = "Penguin bill length (in mm)")

penguins |> 
  ggplot(aes(x = bill_length_mm, fill = species)) + 
  geom_histogram(binwidth = 3) +
  labs(x = "Penguin bill length (in mm)")
```



```{r}
#| warning: false
penguins |> 
  ggplot(aes(x = bill_length_mm, fill = species)) + 
  geom_histogram(binwidth = 3, color = "black") +
  facet_wrap(vars(species), ncol = 1) + 
  theme_bw() + 
  labs(x = "Penguin bill length (in mm)",
       fill = "Species")
```

## Boxplots via `geom_boxplot()`

```{r}
#| warning: false
penguins |> 
  ggplot(aes(x = species, y = bill_length_mm, fill = species)) + 
  geom_boxplot() + 
  labs(y = "Penguin bill length (in mm)")
# Note that the legend is not needed
penguins |> 
  ggplot(aes(x = species, y = bill_length_mm, fill = species)) + 
  geom_boxplot() + 
  guides(fill = "none") + 
  labs(y = "Penguin bill length (in mm)",
       x = "")
```

## Barplots

* `geom_bar()`

```{r}
penguins |> 
  ggplot(aes(x = species)) +
  geom_bar() + 
  labs(x = "")
#
penguins |> 
  ggplot(aes(x = species, fill = species)) +
  geom_bar() + 
  guides(fill = "none") + 
  labs(x = "")
```

