---
title: "Bootstrap Percentile Confidence Interval Problems"
author: "Alan T. Arnholt"
date: 'Last updated: `r format(Sys.time(), "%B %d, %Y at %X")`'
output: bookdown::html_document2
---

```{r, label = "SETUP", echo = FALSE, results= 'hide', message = FALSE, warning = FALSE}
set.seed(123)
library(knitr)
library(tidyverse)
library(PASWR2)
library(infer)
knitr::opts_chunk$set(comment = NA,  fig.align = 'center', fig.height = 5, fig.width = 5, prompt = FALSE, highlight = TRUE, tidy = FALSE, warning = FALSE, message = FALSE)
```

# Bootstrap Distributions

1. Construct a 94% bootstrap percentile confidence interval for the mean number of obstuctive contacts (`oc`) in the `EPIDURAL` data set from the `PASWR2` package.  Use a `for` loop and the `infer` pipeline to construct your confidence intervals.  

```{r}
library(PASWR2)
library(tidyverse)
library(infer)
summary(EPIDURAL)
EPIDURALC <- na.omit(EPIDURAL) # removing the NA value
summary(EPIDURALC)
```
```{r}
B <- 10^4
bsm <- numeric(B)
for(i in 1:B){
  bss <- sample(EPIDURALC$oc, size = sum(!is.na(EPIDURALC$oc)), replace = TRUE)
  bsm[i] <- mean(bss)
}
(quantile(bsm, probs = c(0.03, 0.97)) -> BSPCI)

```

```{r}
# Infer pipeline
EPIDURALC %>% 
  specify(response = oc) %>% 
  generate(reps = B, type = "bootstrap") %>% 
  calculate(stat = "mean") -> bsmip
head(bsmip)
(get_confidence_interval(bsmip, level = 0.94) -> CI1)
# The following computes the same thing
(quantile(bsmip$stat, probs = c(0.03, 0.97)) -> CI2)
```
**The 94% bootsrap percentile confidence interval for the mean number of obstructive
 contacts is: $\text{CI}_{0.94}(\mu_{oc}) = [`r CI1$lower_ci`, `r CI1$upper_ci`]$.**
 
 ___________________


## Visualize the Bootstrap Distribution

2. Create a histogram of the bootstrap distribution of the mean number of obstructive contacts.

```{r}
# Using base R with `bsm`
hist(bsm, xlab = substitute(paste(bar(X),"*")), main = "", col = "gray")
# Using ggplot2 with `bsm`
ggplot(data = data.frame(bsm = bsm), aes(x = bsm)) + 
  geom_histogram(binwidth = 0.1, fill = "gray", color = "black") + 
  theme_bw() + 
  labs(x = substitute(paste(bar(X),"*")))
```

```{r}
# Using wrapper functions with bsmip (an infer object)
visualize(bsmip)
visualize(bsmip, fill = "purple") + 
  labs(x = substitute(paste(bar(X),"*")), 
       title = "Simulation-Based Bootstrap Distribution \n of the Mean Number of Obstructive Contacts") + 
  shade_confidence_interval(CI1, color = "red", fill = "lightblue") + 
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5)) 
```


3. Create a 98% bootstrap percentile confidence interval for the mean `fat` of males in the `BODYFAT` data set from the `PASWR2` package.

```{r}
#Creating a vector with the Male fat values
BODYFAT %>% 
  filter(sex == "M") %>% 
  select(fat) %>% 
  pull() -> fat
fat
# Another approach using base R
subset(BODYFAT, subset = sex == "M", select = fat, drop = TRUE) -> fat2
fat2
# One more approach
BODYFAT$fat[BODYFAT$sex=="M"] -> fat3
fat3
```

```{r}
B <- 10^4
bsmf <- numeric(B)
for(i in 1:B){
  bss <- sample(fat, size = sum(!is.na(fat)), replace = TRUE)
  bsmf[i] <- mean(bss)
}
(quantile(bsmf, probs = c(0.01, 0.99)) -> BSPCIF)
```

```{r}
# Using infer
BODYFAT %>% 
  filter(sex == "M") %>% 
  specify(response = fat) %>% 
  generate(reps = B, type = "bootstrap") %>% 
  calculate(stat = "mean") -> IFM
(get_confidence_interval(IFM, level = 0.98) -> CIF1)
```

## Visualize the Bootstrap Distribution 

```{r}
visualize(IFM) +
  shade_confidence_interval(endpoints = CIF1)
```

_____________

# Bootsrap Distribution for the Difference of Two Means

4. Construct a 90% bootstrap percentile confidence interval for the difference between the mean obstructive contacts using the hamstring stretch position and the mean obstructive contacts using the traditional sitting position.  (Use the `EPIDURAL` data set from `PASWR2`)

```{r}
EPIDURALC %>% 
  filter(treatment == "Hamstring Stretch") %>% 
  select(oc)  %>% 
  pull() -> ochs # OCs for hamstring stretch
EPIDURALC %>% 
  filter(treatment == "Traditional Sitting") %>% 
  select(oc) %>% 
  pull() -> octs # OCs for traditional sitting
####
B <- 10^4
dm <- numeric(B)
for(i in 1:B){

  bsshs <- sample(ochs, size = sum(!is.na(ochs)), replace = TRUE)
  bssts <- sample(octs, size = sum(!is.na(octs)), replace = TRUE)
  dm[i] <- mean(bsshs) - mean(bssts)
}
(quantile(dm, probs = c(0.05, 0.95)) -> CIDM)
```

**The 90% bootstrap percentile confidence interval for the difference in mean obstructive contacts for the hamstring stretch treatment minus the mean obstructive contacts for the traditional sitting position is: $\text{CI}_{0.90}(\mu_{hs} - \mu _{ts})= [`r CIDM[1]`, `r CIDM[2]`]$.**


```{r}
EPIDURALC %>% 
  specify(oc ~ treatment) %>% 
  generate(reps = B, type = "bootstrap") %>% 
  calculate(stat = "diff in means", order = c("Hamstring Stretch", "Traditional Sitting")) -> BSDM
(get_confidence_interval(BSDM, level = 0.90) -> CIDM)
```

## Visualize the Bootstrap Distribution 

```{r}
visualize(BSDM) +
  shade_confidence_interval(endpoints = CIDM)
```


______________

# Bootstrap Percentile Confidence Interval for the Proportion

5.  Construct a 95% bootsrap percentile confidence interval for the true proportion of women who have no complications while receiving an epidural. 

```{r}
results <- EPIDURAL$complications=="None"
(phat <- mean(EPIDURAL$complications=="None"))
B <- 10^4
phatbs <- numeric(B)
for(i in 1:B){
  bss <- sample(results, size = sum(!is.na(results)), replace = TRUE)
  phatbs[i] <- mean(bss)
}
(quantile(phatbs, probs = c(0.025, 0.975)) -> CIP)
```

```{r}
EPIDURAL %>% 
  mutate(no_complications = (complications == "None")) %>% 
  specify(response = no_complications, success = "TRUE") %>% 
  generate(reps = 10^4, type = "bootstrap") %>% 
  calculate(stat = "prop") -> PS
get_confidence_interval(PS, level = 0.95)
```


# Large Sample CI for $p$

```{r}
(n <- sum(!is.na(results)))
phat*n
(1 - phat)*n
CI <- phat +c(-1, 1)*qnorm(.975)*sqrt(phat*(1-phat)/n)
CI
# OR
library(binom)
binom.confint(x = 76, n = 85, conf.level = 0.95, methods = "all")
```
____________

In Class Problem:  Use the data `BODYFAT` from `PASWR2` to construct a 95% percentile bootstrap confidence interval for $\mu_{female\_fat} - \mu_{male\_fat}$.



```{r, echo = FALSE}
BODYFAT %>% 
  specify(fat ~ sex) %>% 
  generate(reps = 10^4, type = "bootstrap") %>% 
  calculate(stat = "diff in means", order = c("F", "M")) -> SBS
quantile(SBS$stat, probs = c(0.025, 0.975), na.rm= TRUE) -> CI1
####
BODYFAT %>% 
  filter(sex == "F") %>% 
  select(fat) %>% 
  pull() -> ffat
BODYFAT %>% 
  filter(sex == "M") %>% 
  select(fat) %>% 
  pull() -> mfat
B <- 10^4
diffmeans <- numeric(B)
for(i in 1:B){
  bssf <- sample(ffat, size = length(ffat), replace = TRUE)
  bssm <- sample(mfat, size = length(mfat), replace = TRUE)
  diffmeans[i] <- mean(bssf) - mean(bssm)
}
quantile(diffmeans, probs = c(0.025, 0.975)) -> CI2
```




