---
title: "Credit Problem"
author: "Alan Arnholt"
date: 'Last updated: `r format(Sys.time(), "%b %d, %Y")`'
output:
  bookdown::html_document2:
    highlight: espresso
    theme: united
---

```{r, label = "SETUP", echo = FALSE, results= 'hide', message = FALSE, warning = FALSE}
set.seed(123)
library(knitr)
knitr::opts_chunk$set(comment = NA, fig.show = 'as.is', fig.align = 'center', fig.height = 5, fig.width = 5, prompt = FALSE, highlight = TRUE, tidy = FALSE, warning = FALSE, message = FALSE, tidy.opts=list(blank = TRUE, width.cutoff= 75, cache = TRUE))
```

* Create a model that predicts an individuals credit rating (`Rating`).

Read in the data:

```{r}
Credit <- read.csv("http://www-bcf.usc.edu/~gareth/ISL/Credit.csv")
Credit$Utilization <- Credit$Balance / (Credit$Income*100)
# DT::datatable(Credit[, -1], rownames = FALSE)
```


```{r, results = "hide"}
library(MASS)
mod.fs <- stepAIC(lm(Rating ~ 1, data = Credit), scope = .~Income + Limit + Cards + Age + Education + Gender + Student + Married + Ethnicity + Balance + Utilization, direction = "forward", test = "F")
mod.be <- stepAIC(lm(Rating ~ Income + Limit + Cards + Age + Education + Gender + Student + Married + Ethnicity + Balance + Utilization, data = Credit), direction = "backward", test = "F")
```

```{r}
summary(mod.fs)
summary(mod.be) 
car::vif(mod.be)
car::vif(mod.fs)
```

I prefer `mod.fs` since it has less multicollinearity.

* Create another model that predicts rating with `Limit`, `Cards`, `Married`, `Student`, and `Education` as features. 

```{r}
mod <- lm(Rating ~ Limit + Cards + Married + Student + Education, data = Credit)
summary(mod)
par(mfrow = c(2, 2))
plot(mod)
par(mfrow = c(1, 1))
```

* Use your model to predict the `Rating` for an individual that has a credit card limit of
$6,000, has 4 credit cards, is married, and is not a student, and has an undergraduate degree (`Education` = 16).

```{r}
predict(mod, newdata = data.frame(Limit = 6000, Cards = 4, Married = "Yes", Student = "No", Education = 16), response = "pred")
```

* Use your model to predict the `Rating` for an individual that has a credit card limit of
$12,000, has 2 credit cards, is married, is not a student, and has an eighth grade education (`Education` = 8).

```{r}
predict(mod, newdata = data.frame(Limit = 12000, Cards = 2, Married = "Yes", Student = "No", Education = 8), response = "pred")
```

