---
title: "Correlation and Regression"
author: "Alan T. Arnholt"
date: 'Last updated: `r format(Sys.time(), "%B %d, %Y at %X")`'
output: bookdown::html_document2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = NA, warning = FALSE, 
                      message = FALSE, fig.align = "center")
```

# Correlation

## Some Data

```{r}
set.seed(13)
x <- 1:20
y <- x + rnorm(20, 0, 3)
DFP <- data.frame(x = x, y = y)
y <- x + rnorm(20, 0, 3)
DFN <- data.frame(x = -x, y = y)
rm(x, y)
library(tidyverse)
ggplot(data = DFP, aes(x = x, y = y)) + 
  geom_point(color = "blue") + 
  theme_bw()
ggplot(data = DFN, aes(x = x, y = y)) + 
  geom_point(color = "blue") + 
  theme_bw()

```


$$r = \dfrac{1}{n-1}\sum_{i=1}^n\frac{(x - \bar{x})}{s_x}\frac{(y - \bar{y})}{s_y}$$

```{r}
DFP %>% 
  mutate(x_xbar = x - mean(x), y_ybar = y - mean(y), zx = x_xbar/sd(x), zy = y_ybar/sd(y), zxzy = zx*zy) -> steps
steps
steps %>% 
  summarize(COR = sum(zxzy)/(length(x) -1), cor(x,y))
```

```{r}
DFN %>% 
  summarize(r = cor(x, y))
```

## Least Squares Regression


$$b_1 = \frac{\sum_{i=1}^{n}(x_i - \bar{x})(y_i - \bar{y})}{\sum_{i=1}^{n}(x_i - \bar{x})^2} = r\cdot \frac{s_y}{s_x}$$

$$b_0 = \bar{y} - b_1\cdot\bar{x}$$

$$e_i = y_i - \hat{y_i}$$

$$\hat{y_i} = b_0 + b_1\cdot x_i$$

```{r}
DFP %>% 
  mutate(b1 = cor(x, y)*sd(y)/sd(x), b0 = mean(y)-b1*mean(x), yhat = b0 + b1*x, e = y - yhat) -> ans1
ans1
ans1 %>% 
  summarize(SSE = sum(e^2))
```

```{r}
mod <- lm(y ~ x, data = DFP)
summary(mod)
anova(mod)
anova(mod)[2, 2] -> SSE
SSE
anova(mod)[2, 3] -> MSE
MSE
RSE <- sqrt(MSE)
RSE
summary(mod)$sigma
```

## Adding least squares line to scatterplot

```{r}
ggplot(data = DFP, aes(x = x, y = y)) + 
  geom_point(color = "lightblue") + 
  theme_bw() + 
  geom_smooth(method = "lm", se = FALSE, color = "red")
```

```{r}
summary(mod)
coefficients(summary(mod)) -> STUFF
STUFF
b0 <- STUFF[1, 1]
b1 <- STUFF[2, 1]
c(b0, b1)
coef(mod)
```

## Using `moderndive` wrapper functions

* `get_regression_table()`
* `get_regression_points()`


```{r}
library(moderndive)
get_regression_table(mod) -> OUT 
OUT 
OUT$estimate
```

```{r}
get_regression_points(mod)
```

## Prediction

What does the `mod` predict for an `x` value of 15?

```{r}
yhat <- b0 + b1*15
yhat
# or
predict(mod, newdata = data.frame(x = 15))
#
predict(mod, newdata = data.frame(x = 15), interval = "conf", level = 0.95)
```

```{r}
ggplot(data = DFP, aes(x = x, y = y)) + 
  theme_bw() + 
  geom_smooth(method = "lm", color = "lightblue") + 
  geom_point(color = "purple")
```

## You try

Use the data set `VIT2005` from the `PASWR2` package to regress `totalprice` onto `area` and store the resulting model in an object named `mod2`.  Use `mod2` to predict the price of an apartment that has 100 square meters.

```{r}
# Your code here
```

