---
title: "`infer` Examples"
author: "Notes"
date: '`r Sys.Date()`'
output: bookdown::html_document2
---

```{r, label = "SETUP", echo = FALSE, results= 'hide', message = FALSE, warning = FALSE}
library(knitr)
library(tidyverse)
knitr::opts_chunk$set(comment = NA, fig.show = 'as.is', fig.align = "center", fig.height = 4.5, fig.width = 5.5, prompt = FALSE, highlight = TRUE, tidy = FALSE, warning = FALSE, message = FALSE, tidy.opts=list(blank = TRUE, width.cutoff= 75, cache = TRUE))
```

# Various pipelines

To read about the `mtcars` data set, type `?mtcars` at the `R` prompt.

```{r}
str(mtcars)
DT::datatable(mtcars)
```

We want to test $H_0: \mu_{\text{Manual}} - \mu_{\text{Automatic}} = 0$ versus $H_A: \mu_{\text{Manual}} - \mu_{\text{Automatic}} > 0$


To do so, first change the variable `am` to a factor and reorder the levels of the variable.  

```{r}
mtcars <- mtcars %>% 
  mutate(am = factor(am, levels = c("1","0"), labels = c("Manual", "Automatic")),
         cyl = factor(cyl, levels = c("4", "6", "8")))
mtcars %>% 
  group_by(am) %>% 
  summarize(Mean = mean(mpg), SD = sd(mpg), n = n())
mtcars %>% 
  group_by(cyl) %>% 
  summarize(Mean = mean(mpg), SD = sd(mpg), n = n())
```
```{r, fig.width = 10}
p1 <- ggplot(data = mtcars, aes(x = mpg)) +
  geom_histogram(binwidth = 3, fill = "lightblue", color = "black") +
  facet_wrap(~am) + 
  theme_bw()
#
p2 <- ggplot(data = mtcars, aes(x = mpg)) +
  geom_density(fill = "lightblue", color = "black") +
  facet_wrap(~am) + 
  theme_bw()
p3 <- ggplot(data = mtcars, aes(sample = mpg, color = am)) + 
  geom_qq() + 
  facet_wrap(.~am) + 
  theme_bw()
#
p4 <- ggplot(data = mtcars, aes(sample = mpg, color = am)) + 
  geom_qq() + 
  theme_bw() + 
  scale_color_manual(values = c("blue", "red"))
gridExtra::grid.arrange(p1, p2, p3, p4, ncol = 2)
```


```{r}
SIMS <- 1000  # set this 
library(infer)
mtcars %>%
    specify(mpg ~ am) %>% # alt: response = mpg, explanatory = am
    hypothesize(null = "independence") %>%
    generate(reps = SIMS, type = "permute") %>%
    calculate(stat = "t", order = c("Manual", "Automatic")) %>%  # 0 = automatic, 1 = manual
    visualize(method = "both") #"simulation" is the default method
```

```{r}
mtcars %>%
    specify(response = mpg, explanatory = am) %>% # alt: mpg ~ am
    hypothesize(null = "independence") %>%
    generate(reps = SIMS, type = "permute") %>%
    calculate(stat = "t", order = c("Manual", "Automatic")) %>%  # 0 = automatic, 1 = manual
    visualize(method = "both") #"simulation" is the default method
```
```{r}
TS <- mtcars %>%
    specify(mpg ~ am) %>% # alt: response = mpg, explanatory = am
    hypothesize(null = "independence") %>%
    generate(reps = SIMS, type = "permute") %>%
    calculate(stat = "t", order = c("Manual", "Automatic")) # 0 = automatic, 1 = manual
#####
obs_stat <- t.test(mpg ~ am, data = mtcars)$stat
obs_stat
####
ggplot(data =TS, aes(x = stat, y = ..density..)) +
  geom_histogram(fill = "lightblue", color = "black", binwidth = 0.3) + 
  theme_bw() + 
  geom_vline(xintercept = obs_stat, linetype = "dashed") 
####
get_pvalue(TS, obs_stat, direction = "greater")
####
pvalue <- (sum(TS$stat >= obs_stat) + 1)/(SIMS + 1)
pvalue
```

## Using a `for()` loop

```{r}
sims <- SIMS
ts <- numeric(sims)
for(i in 1:sims){
  ts[i] <- t.test(mpg ~ sample(am), data = mtcars)$stat
}
pvalue <- (sum(ts >= obs_stat) + 1)/(sims + 1)
pvalue
```

## Using Difference in Means

```{r}
DM <- mtcars %>%
    specify(mpg ~ am) %>% # alt: response = mpg, explanatory = am
    hypothesize(null = "independence") %>%
    generate(reps = SIMS, type = "permute") %>%
    calculate(stat = "diff in means", order = c("Manual", "Automatic")) # 0 = automatic, 1 = manual
#####
obs_MDIFF <- mtcars %>% 
  group_by(am) %>% 
  summarize(MD = mean(mpg)) %>% 
  mutate(MeanDiff = -diff(MD)) %>% 
  pull()
obs_stat <- obs_MDIFF[1]
obs_stat
get_pvalue(DM, obs_stat = obs_stat, direction = "greater")
pvalue <- mean(DM$stat >= obs_stat)
pvalue
# Better yet
pvalue1 <- (sum(DM$stat >= obs_stat) + 1)/(SIMS + 1)
pvalue1
```

### With a `for` loop

```{r}
sims <- SIMS - 1
dm <- numeric(sims)
for(i in 1:sims){
  dm[i] <- -diff(tapply(mtcars$mpg, sample(mtcars$am), mean))
}
ggplot(data = data.frame(stat = dm), aes(x = stat)) + 
  geom_density(fill = "purple") + 
  theme_bw() + 
  geom_vline(xintercept = obs_stat, linetype = "dashed")
pvalue <- (sum(dm >= obs_stat) + 1)/(sims + 1)
pvalue
```

## Confidence Intervals

Consturct a 92% CI for the true mean quarter mile time.

```{r}
BM <- specify(mtcars, response = qsec) %>% 
  generate(reps = SIMS, type = "bootstrap") %>% 
  calculate(stat = "mean")
visualize(BM)
```

```{r}
# Bootstrap Percentile CI
PCI <- BM %>%
  summarize(l = quantile(stat, 0.04),
            u = quantile(stat, 0.96)) 
PCI
###
# Visualize
BM %>% visualize(endpoints = PCI, 
                          direction = "between",
                          ci_fill = "pink") + 
  theme_bw()
####
ggplot(data = BM, aes(x = stat, fill = stat > PCI$l & stat < PCI$u) ) +
  geom_histogram(binwidth = 0.15) + 
  scale_fill_manual(values = c("blue", "green")) +
  theme_bw()

```

```{r}
# Bootstrap SE CI
(n <- sum(!is.na(mtcars$qsec)))
(multiplier <- qt(0.96, n - 1))
(SEBM <- sd(BM$stat))
(CI <- mean(mtcars$qsec) + c(-1, 1)*multiplier*SEBM)
#
# Visualize
BM %>% visualize(endpoints = CI, 
                          direction = "between",
                          ci_fill = "yellow") + 
  theme_bw()
```

### Using a `for()` loop for CI

```{r}
sims <- SIMS
ME <- numeric(sims)
for(i in 1:sims){
  bss <- sample(mtcars$qsec, size = sum(!is.na(mtcars$qsec)), replace = TRUE)
  ME[i] <- mean(bss)
}
# Percentile CI
(PCI <- quantile(ME, probs = c(0.04, 0.96)))
# SE CI
(multiplier <- qt(0.96, 32 - 1))
(SCI <- mean(mtcars$qsec) + c(-1, 1)*multiplier*sd((ME)))
```

## ANOVA

```{r}
mtcars %>% 
  group_by(cyl) %>% 
  summarize(Mean = mean(mpg), SD = sd(mpg), n = n())

(GM <- mean(mtcars$mpg))
(MT1 <- mean(mtcars$mpg[mtcars$cyl == "4"]))
(MT2 <- mean(mtcars$mpg[mtcars$cyl == "6"]))
(MT3 <- mean(mtcars$mpg[mtcars$cyl == "8"]))

(SSTreat <- 11*(MT1 - GM)^2 + 7*(MT2 - GM)^2 + 14*(MT3 - GM)^2)
(SSTotal <- sum((mtcars$mpg - GM)^2))
(SSError <- SSTotal - SSTreat)
```

```{r}
# Checks
ggplot(data = mtcars, aes(sample = mpg)) +
  geom_qq() +
  facet_wrap(.~cyl) + 
  theme_bw()
# Questionable whether the assumption of constant variance across treatments
# is satisfied
ggplot(data = mtcars, aes(x = cyl, y = mpg)) +
  geom_boxplot() + 
  theme_bw()
# If assumptions are satisfied use what follows
mod.aov <- aov(mpg ~ cyl, data = mtcars)
summary(mod.aov)
obs_stat <- summary(mod.aov)[[1]][1, 4]
obs_stat
# OR
obs_stat <- anova(lm(mpg~cyl, data = mtcars))[1, 4]
obs_stat
```

## Using a resampling approach to test equality of multiple means

$H_0:\mu_4 = \mu_6 = \mu_8$ versus $H_A: \mu_i \neq \mu_j \text{ for some } (i, j)$

### `infer` approach

```{r}
FS <- mtcars %>% 
  specify(mpg ~ cyl) %>% 
  hypothesize(null = "independence") %>% 
  generate(reps = SIMS, type = "permute") %>% 
  calculate(stat = "F")
visualize(FS, method = "both")
get_pvalue(FS, obs_stat = obs_stat, direction = "greater")
```

### `for()` loop

```{r}
sims <- SIMS -1
FFS <- numeric(sims)
for(i in 1:sims){
  FFS[i] <- summary(aov(mpg ~ sample(cyl), data = mtcars))[[1]][1, 4]
}
pvalue <- (sum(FFS >= obs_stat) + 1)/(sims + 1)
pvalue
```

```{r}
ggplot(data = data.frame(stat = FFS), aes(x = stat)) +
  geom_density(color = "red", fill = "tomato") + 
  theme_bw() + 
  stat_function(fun = df, args = list(2, 29), color = "blue")
```


## Testing a single parameter with a bootstrap approach

Consider testing $H_0:\mu_{qsec} = 17$ versus $H_A:\mu_{qsec} \neq 17$

```{r}
(xbar <- mean(mtcars$qsec))
(null <- 17 - xbar)

sims <- SIMS
bm <- numeric(sims)
for(i in 1:sims){
  bss <- sample(mtcars$qsec, size = 32, replace = TRUE) + null
  bm[i] <- mean(bss)
}
# Base histogram
hist(bm, breaks = "Scott", col = "lightblue")
# ggplot2 histogram
ggplot(data = data.frame(x = bm), aes(x = x)) +
  geom_histogram(color = "black", fill = "lightblue", binwidth = 0.1) + theme_bw()
  
pvalue <- (sum(bm >= xbar)*2 + 1)/(sims + 1)
pvalue
# Or
pvalue2 <- (sum(bm >= 17 - null) + sum(bm <= 17 + null) + 1)/(sims + 1)
pvalue2
#
t.test(mtcars$qsec, mu = 17, alternative = "two.sided")
```


