---
title: "R Markdown"
author: "Alan T. Arnholt"
date: '`r format(Sys.time(), "%b %d, %Y")`'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = NA, fig.align = "center", message = FALSE)
```

## Understanding the data

* Use `summary()` and `str()`

* Use the short cut to insert a code chunk! - `Ctrl +  Alt  + I`

```{r}
summary(UCBAdmissions)
str(UCBAdmissions)
```

## Wrangle data

### Using `dplyr`

Main verbs:

* Pick observations by their values with `filter()`
* Reorder the rows with `arrange()`
* Pick variables by their name with `select()`
* Create new variables with `mutate()`
* Collapse many values to a single summary with `summarize()`

See the [data transformation cheat sheet](https://github.com/rstudio/cheatsheets/raw/master/source/pdfs/data-transformation-cheatsheet.pdf) for more information.

```{r}
UCB <- as.data.frame(UCBAdmissions)
UCB
# Not quite what we want yet
library(dplyr)
ucb_admit <- UCB %>% 
  slice(rep(1:n(), Freq)) %>% 
  select(-Freq)
ucb_admit
```

## A few summary numbers

```{r}
ucb_admit %>%
  count(Admit, Gender)
```

Does this look like discrimination?  We will display the data using `ggplot2`.  First we will need to load the package using `library()`.  See the [Data Visualization with `ggplot2` cheat sheet](https://github.com/rstudio/cheatsheets/raw/master/source/pdfs/ggplot2-cheatsheet-2.1.pdf) for more details.

```{r}
library(ggplot2)
ggplot(data = ucb_admit, aes(x = Gender, fill = Admit)) + 
  geom_bar(position = "dodge")
```

What do you think?

```{r}
ggplot(data = ucb_admit, aes(x = Gender, fill = Admit)) + 
  geom_bar(position = "fill") + 
  labs(y = "Fraction") +
  scale_fill_manual(values = c("green", "red")) + 
  theme_bw()
```

## Computing the actual numbers

* To insert the pipe operator (`%>%`) use the short cut `Ctrl + Shift + M`

```{r}
library(tidyr)
ucb_admit %>%
  count(Admit, Gender) 
ucb_admit %>%
  count(Admit, Gender) %>%
  spread(Admit, n)
ucb_admit %>%
  count(Admit, Gender) %>%
  spread(Admit, n) %>% 
  mutate(Percent_Rejected = Rejected / (Admitted + Rejected),
         Percent_Accepted = Admitted /(Admitted + Rejected))
```

Next we store the `tibble` as the object `basicUCB`.

```{r}
basicUCB <- ucb_admit %>%
  count(Admit, Gender) %>%
  spread(Admit, n) %>% 
  mutate(Percent_Rejected = Rejected / (Admitted + Rejected),
         Percent_Accepted = Admitted /(Admitted + Rejected))
basicUCB
```

A graph of the percent that were accepted follows.

```{r}
ggplot(data = basicUCB, aes(x = Gender, y = Percent_Accepted)) + 
  geom_bar(stat = "identity", fill = c("blue", "pink")) + 
  theme_bw()
```

Next, we look at the percentages in each department.

```{r}
UCBres <- ucb_admit %>%
  group_by(Dept) %>%
  count(Admit, Gender) %>%
  spread(Admit, n) %>%
  mutate(Percent_Accepted = Admitted/(Admitted + Rejected), 
         Percent_Rejected = Rejected/(Admitted + Rejected))
UCBres
```

Graphing the information in `UCBres`.

```{r}
ggplot(data = UCBres, aes(x = Dept, y = Percent_Accepted, fill = Gender)) + 
  geom_bar(stat = "identity", position = "dodge") + 
  labs(y = "Percent Admitted", x = "Department") + 
  theme_bw() + 
  scale_fill_manual(values = c("blue", "pink"))
```

```{r}
ggplot(data = UCBres, aes(x = Dept, y = Percent_Rejected, fill = Gender)) + 
  geom_bar(stat = "identity", position = "dodge") + 
  labs(y = "Percent Rejected", x = "Department") + 
  theme_bw() + 
  scale_fill_manual(values = c("blue", "pink"))
```

## Visualizing Categorical Data

* Use the `vcd` package

```{r}
library(vcd)
mosaic(~Gender, data = UCB)
mosaic(~Gender + Admit, data = UCB)
mosaic(~Gender + Admit, data = UCB, shade = TRUE)
mosaic(~Dept + Gender + Admit, data = UCB, shade = TRUE)
mosaic(~Dept + Gender + Admit, data = UCB, shade = TRUE, 
       direction = c("v", "h", "v"))
mosaic(~Dept + Gender + Admit, data = UCB, shade = TRUE, 
       direction = c("v", "h", "v"), highlighting = "Admit")
```


