---
title: "Chi-Square"
author: "Notes"
date: 'Last compiled: `r format(Sys.time(), "%A, %B %d, %Y - %X.")`'
output: bookdown::html_document2
---

```{r, label = "SETUP", echo = FALSE, results= 'hide', message = FALSE, warning = FALSE}
library(knitr)
knitr::opts_chunk$set(comment = NA, fig.show = 'as.is', fig.align = "center", fig.height = 4.5, fig.width = 5.5, prompt = FALSE, highlight = TRUE, tidy = FALSE, warning = FALSE, message = FALSE, tidy.opts=list(blank = TRUE, width.cutoff= 75, cache = TRUE))
library(tidyverse)
library(infer)
library(openintro)
library(PASWR2)
# Parallel Processing
library(doMC)
registerDoMC(cores = 12)
```

Is there a relationship between `pclass` and `survived`?

```{r}
TITANIC3 <- TITANIC3 %>% 
  mutate(survived = factor(survived, labels = c("Dead", "Alive")))
ggplot(data = TITANIC3, aes(x = survived, fill = pclass)) +
  geom_bar(position = "fill") +
  theme_bw() + 
  labs(y = "percent")
```


# Chi-Square

**Note:** $E_{ij} = \frac{R_i \times C_j}{n_T}$

$$\chi^2_{obs} = \sum\frac{(O- E)^2}{E}$$

```{r}
T1 <- TITANIC3 %>% 
  select(pclass, survived) %>% 
  table() 
T1
addmargins(T1)
E11 <- (323/1309)*(809/1309)*1309
E12 <- (323/1309)*(500/1309)*1309
E21 <- (277/1309)*(809/1309)*1309
E22 <- (277/1309)*(500/1309)*1309
E31 <- (709/1309)*(809/1309)*1309
E32 <- (709/1309)*(500/1309)*1309
EIJ <- matrix(c(E11, E12, E21, E22, E31, E32), byrow = TRUE, nrow = 3)
EIJ
```

```{r}
X2 <- (123 - E11)^2/E11 + (200 - E12)^2/E12 + (158 - E21)^2/E21 + (119 - E22)^2/E22 +
      (528 - E31)^2/E31 + (181 - E32)^2/E32 
X2
pvalue <- pchisq(X2, 2, lower = FALSE)
pvalue
chisq.test(T1, correct = FALSE)
# Easier fashion to get expected
chisq.test(T1)$exp
```


## Same thing with permutation

### Using a `for()` loop

```{r}
B <- 10^4 - 1
X2 <- numeric(B)
for(i in 1:B){
  PT <- xtabs(~pclass + sample(survived), data = TITANIC3)
  X2[i] <- chisq.test(PT, correct = FALSE)$stat
}
chi_obs_stat <- chisq.test(xtabs(~pclass + survived, data = TITANIC3))$stat
chi_obs_stat
# OR
chi_obs_stat <- chisq.test(table(TITANIC3$pclass, TITANIC3$survived))$stat
chi_obs_stat
#
pvalueSIM <- (sum(X2 >= chi_obs_stat) + 1)/(B + 1)
pvalueSIM
```

What does the permutation distribution under the null hypothesis look like?

```{r}
ggplot(data = data.frame(x = X2), aes(x = x)) +
  geom_density(fill = "pink") + 
  theme_bw() 
```

## Same thing with `infer`

```{r}
null <- TITANIC3 %>% 
  specify(pclass ~ survived) %>% 
  hypothesize(null = "independence") %>% 
  generate(reps = 10^3 - 1, type = "permute") %>% 
  calculate(stat = "Chisq")
```

### Visualize the null distribution

```{r}
ggplot(data = null, aes(x = stat)) +
  geom_density(fill = "skyblue") +
  theme_bw()
```

```{r}
visualize(null, method = "both", dens_color = "blue") + 
  theme_bw()
```


```{r}
null %>% 
  summarize(pvalue = (sum(stat >= chi_obs_stat) + 1)/(10^3)) 
```

**Note problem below**

```{r}
get_p_value(null, obs_stat = chi_obs_stat, direction = "right")
```


# Consider testing $H_0:p_1 - p_2 = 0$



```{r}
TITANIC3 %>% 
  select(sex, survived) %>% 
  table() %>% 
  addmargins()
TITANIC3 %>% 
  select(sex, survived) %>% 
  table() %>% 
  prop.table(1)
```

Consider testing: $H_0:p_{\text{Male Alive}} - p_{\text{Female Alive}} = 0$ versus $H_A:p_{\text{Male Alive}} - p_{\text{Female Alive}} < 0$

```{r}
prop.test(x = c(161, 339), n = c(843, 466), alternative = "less", correct = FALSE)
obs_stat <- prop.test(x = c(161, 339), n = c(843, 466), alternative = "less", correct = FALSE)$stat^.5
obs_stat
# Same as
phatmale <- 161/843
phatfemale <- 339/466
phatp <- (161 + 339)/(843 + 466)
zobs <- (phatmale - phatfemale) / sqrt( phatp*(1 - phatp)/843 + phatp*(1 - phatp)/466)
zobs
pvalue <- pnorm(zobs)
pvalue
```

Note: This can also be tested using the $\chi^2$-test of homogeneity.

$H_0: p_{\text{Male Alive}} = p_{\text{Female Alive}} = p_{\text{Male Dead}} = p_{\text{Female Dead}}$ versus at least some $p_{ij} \neq p_{i+1, j}$ for some $(i,j)$.

```{r}
T1 <- TITANIC3 %>% 
  select(sex, survived) %>% 
  table()
chisq.test(T1, correct = FALSE)
chisq.test(T1, correct = FALSE)$stat^.5
```

```{r}
null <- TITANIC3 %>% 
  specify(survived ~ sex, success = "Alive") %>% 
  hypothesize(null = "independence") %>% 
  generate(reps = 10^3 - 1, type = "permute") %>% 
  calculate(stat = "Chisq", order = c("male", "female"))
ggplot(data = null, aes(x  = stat)) + 
  geom_density(fill = "pink") +
  theme_bw()
null %>%
  summarize( (sum(stat >= obs_stat) + 1)/10^3)
```

```{r}
B <- 10^4 - 1
X2 <- numeric(B)
for(i in 1:B){
  T3 <- xtabs(~sex + sample(survived), data = TITANIC3)
  X2[i] <- chisq.test(T3, correct = FALSE)$stat 
}
pvalue <- (sum(X2 >= obs_stat) + 1)/10^4
pvalue
```

# Chi-Square test of Homogeneity

Data will often come summarized in contingency tables.  

```{r}
DP <- c(67, 76, 57, 48, 73, 79)
MDP <- matrix(data = DP, nrow = 2, byrow = TRUE)
dimnames(MDP) <- list(Pop = c("Drug", "Placebo"), Status = c("Improve", "No Change", "Worse"))
TDP <- as.table(MDP)
TDP
```

The problem then becomes one of putting the data back into a `tidy` format.

```{r}
NT <- TDP %>% 
  tibble::as_tibble() %>% 
  uncount(n)
NT
```


## Graph

```{r}
ggplot(data = NT, aes(x = Pop, fill = Status)) +
  geom_bar(position = "fill") +
  theme_bw()
```



```{r}
T1 <- NT %>% 
  table()
T1
chisq.test(T1)
(chi_obs <- chisq.test(T1)$stat)
E <- chisq.test(T1)$exp
E
(X2 <- sum((T1 - E)^2/E))
(pvalue <- pchisq(X2, 2, lower = FALSE))
```

## Permutation distribution

```{r}
B <- 10^3 - 1
X2 <- numeric(B)
for(i in 1:B){
  TN <- xtabs(~Pop + sample(Status), data = NT)
  X2[i] <- chisq.test(TN)$stat
}
null_hom <- data.frame(stat = X2)
ggplot(data = null_hom, aes(x = stat)) +
  geom_density(fill = "purple", alpha = 0.3) +
  theme_bw() + 
  stat_function(fun = dchisq, args = list(df = 2), color = "blue") 
(pval <- (sum(null_hom$stat >= chi_obs) + 1)/(B + 1))
```

## Same approach with `infer`

```{r}
null_hom_infer <- NT %>% 
  specify(Pop ~ Status) %>% 
  hypothesize(null = "independence") %>% 
  generate(reps = 10^3 - 1, type = "permute") %>% 
  calculate(stat = "Chisq")
visualize(null_hom_infer, method = "both", dens_color = "blue") +
  theme_bw()
(pval2 <- (sum(null_hom_infer$stat >= chi_obs) + 1)/(B + 1))
```
## GOF test

Given another table (from summarized information):

```{r}
BD <- c(23, 20, 18, 23, 20, 19, 18, 21, 19, 22, 24, 29)
Sign = c("Aries", "Taurus", "Gemini", "Cancer", "Leo",
         "Virgo", "Libra", "Scorpio", "Sagitarius",
         "Capricorn", "Aquarius", "Pisces")
VEC <- rep(Sign, times = BD)
(T2 <- table(VEC))

BNT <- T2 %>% 
  tibble::as_tibble() %>% 
  uncount(n)
BNT
```

## Graph

```{r}
ggplot(data = BNT, aes(x = VEC)) +
  geom_bar(fill = rainbow(12)) +
  theme_bw() + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
  labs(x = "Astrological Sign")
```


```{r}
chisq.test(T2, p = rep(1/12, 12))
(pv <- chisq.test(T2, p = rep(1/12, 12))$p.value)
```
### Conclusion

There is little evidence (p-value = `r pv`) to suggest a nonuniform distribution of zodiac signs among executives.

## Chi-Square Test of Independence

```{r}
TH <- c(17, 35, 8, 53, 22, 491)
MTH <- matrix(data = TH, nrow = 3, byrow = TRUE)
dimnames(MTH) <- list(Tatoo = c("Parlor", "Elsewhere", "None"), HepC = c("Yes", "No"))
MTH <- as.table(MTH)
MTH
NHT <- MTH %>% 
  tibble::as_tibble() %>% 
  uncount(n)
NHT
```

```{r}
ggplot(data = NHT, aes( x = HepC, fill = Tatoo)) +
  geom_bar(position = "fill") +
  theme_bw() +
  scale_fill_manual(values = c("pink", "green", "red"))
```

```{r}
(T1 <- xtabs(~HepC + Tatoo,data = NHT))
chisq.test(T1)
(obs_stat <- chisq.test(T1)$stat)
```

```{r, warning = FALSE}
B <- 10^4 - 1
X2 <- numeric(B)
for(i in 1:B){
  TT <- xtabs(~HepC + sample(Tatoo), data = NHT)
  X2[i] <- chisq.test(TT)$stat
}
(pval <- (sum(X2 >= obs_stat) + 1)/(10^4 -1 + 1))
```

