---
title: "Problem 5.14"
author: "From Class"
date: '`r format(Sys.time(), "%B %d, %Y at %X")`'
output: bookdown::html_document2
bibliography: PackagesUsed.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = NA, fig.align = "center")
library(tidyverse)
library(infer)
library(resampledata)
library(gridExtra)
```


```{r, results = "hide", echo = FALSE}
PackagesUsed <- c("tidyverse", "infer", "resampledata", "gridExtra", "rmarkdown", "bookdown")
# Write bib information
knitr::write_bib(PackagesUsed, file = "./PackagesUsed.bib")
# Load packages
lapply(PackagesUsed, library, character.only = TRUE)
```

This document uses `tidyverse` by @R-tidyverse, `infer` by @R-infer, `resampledata` by @R-resampledata, `gridExtra` by @R-gridExtra, and `bookdown` by @R-bookdown.

_________________

**Directions:** 

1. Add comments to the code to make sure you understand what all steps are doing.
2. Add descriptive titles and labels to the base `R` histograms.
3. Discuss the difference between the permutation distribution and the bootstrap distribution.  Make sure to discuss where each distribution is centered.

_____________________

This problem roughly follows 5.14 from page 145 of your text.  One important difference is that we are performing a two-sided test (not a directional test) as the book requests.

______________________

$$H_O: \mu_M - \mu_F = 0$$
$$H_A: \mu_M - \mu_F \neq 0$$

_________________


```{r}
head(Beerwings)
```

```{r}
ggplot(data = Beerwings, aes(x = Gender, y = Hotwings)) + 
  geom_boxplot() + 
  theme_bw() -> p1
ggplot(data = Beerwings, aes(sample = Hotwings, color = Gender)) + 
  stat_qq() +
  stat_qq_line() +
  theme_bw() -> p2
grid.arrange(p1, p2, nrow = 1)
```

```{r}
Beerwings %>%
  group_by(Gender) %>%
  summarize(Means = mean(Hotwings), SD = sd(Hotwings), n = n())

Beerwings %>%
  group_by(Gender) %>%
  summarize(Means = mean(Hotwings)) %>%
  select(Means) %>%
  pull() %>%
  diff() -> obs_diff
obs_diff
```





```{r}
BWM <- Beerwings$Hotwings[Beerwings$Gender == "M"]
BWF <- Beerwings$Hotwings[Beerwings$Gender == "F"]
obs_diff <- mean(BWM) - mean(BWF)
obs_diff
```

```{r}
Beerwings %>%
  filter(Gender == "M") %>%
  select(Hotwings) %>%
  pull() -> BWM
BWM
Beerwings %>%
  filter(Gender == "F") %>%
  select(Hotwings) %>%
  pull() -> BWF
BWF
obs_diff <- mean(BWM) - mean(BWF)
obs_diff
```

```{r}
tapply(Beerwings$Hotwings, Beerwings$Gender, mean)
diff(tapply(Beerwings$Hotwings, Beerwings$Gender, mean)) -> obs_diff
obs_diff
```

```{r}
HWM <- subset(Beerwings, select = Hotwings, subset = Gender == "M", drop = TRUE)
HWF <- subset(Beerwings, select = Hotwings, subset = Gender == "F", drop = TRUE)
HWM
HWF
c(mean(HWF), mean(HWM))
diff(c(mean(HWF), mean(HWM))) -> obs_diff
obs_diff
```

# Generating the permutation distribution
## Using `infer`

```{r}
Beerwings %>%
  specify(Hotwings ~ Gender) %>%
  hypothesize(null = "independence") %>%
  generate(reps = 10000, type = "permute") %>%
  calculate(stat = "diff in means", order = c("M", "F")) -> permDist
permDist %>%
  visualize() +
  theme_bw() + 
  geom_vline(xintercept = c(-obs_diff, obs_diff))

pvalue <- mean(permDist$stat >= obs_diff)*2
pvalue
#
get_p_value(permDist, obs_stat = obs_diff, direction = "two-sided")
```
## Using `for()`

```{r}
B <- 10000
diff_means <- numeric(B)
for(i in 1:B){
  diff_means[i] <- diff(tapply(Beerwings$Hotwings, sample(Beerwings$Gender), mean))
}
hist(diff_means)
abline(v = c(-obs_diff, obs_diff))
pvalue <- mean(diff_means >= obs_diff)*2
pvalue
```

Based on a _p-value_ of `r pvalue` which is less than $\alpha = 0.05$ we reject the null hypothesis and conclude that there is evidence to suggest the average wings consumed by males is not the same as the average number of wings consumed by females.

__________________

# Constructing 95% Bootstrap Percentile CI

## Using `infer`

```{r}
Beerwings %>% 
  specify(Hotwings ~ Gender) %>% 
  generate(reps = 10000, type = "bootstrap") %>% 
  calculate(stat = "diff in means", order = c("M", "F")) -> bootdist
bootdist %>% 
get_confidence_interval(level = 0.95, type = "percentile") -> PCI
PCI
visualize(bootdist) +
  shade_confidence_interval(endpoints = PCI, fill = "yellow") + 
  theme_bw()
```

We are 95% confident that the true difference between the mean hotwings consumed by males and the mean hotwings consumed by females fall in the interval $(`r round(PCI[1],1)` \text{ hotwings}, `r round(PCI[2],1)` \text{ hotwings})$.

## Using a `for()` loop

```{r}
B <- 10000
diff_means <- numeric(B)
for(i in 1:B){
  bssM <- sample(HWM, size = 15, replace = TRUE)
  bssF <- sample(HWF, size = 15, replace = TRUE)
  diff_means[i] <- mean(bssM) - mean(bssF)
}
quantile(diff_means, probs = c(0.025, 0.975)) -> PCI2
PCI2
hist(diff_means)
abline(v = PCI2)
```

__________

We are 95% confident that the true difference between the mean hotwings consumed by males and the mean hotwings consumed by females fall in the interval $(`r round(PCI2[1],1)` \text{ hotwings}, `r round(PCI2[2],1)` \text{ hotwings})$.

_________

## References