---
title: "Some Probability Problems"
author: "Alan T. Arnholt"
date: "9/30/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, comment = NA)
```

According to the Bureau of the Censes, 50% of all homes in Dallas are heated with gas.  For three randomly selected homes in Dallas, list the outcomes of the sample space corresponding to whether they are heated with gas.  Also list the outcomes that make up the following events:

  * Event A: Exactly two of the three heat with gas
  
  * Event B: Only one heats with gas
  
  * Event C: all three heat with gas
  
  * Event D: At least one of the three heats with gas
  
$\Omega = \{(g,g,g), (g,g,n), (g,n,g), (n,g,g), (g,n,n), (n,g,n), (n,n,g), (n,n,n)\}$ 


Simulate the sample space using 10,000 simulations with the `rbinom()` function.  Find the simulated probabilities for events A, B, C, and D.

```{r}
S <- rbinom(100000, 3, 0.5)
table(S)
# P(A)
mean(S==2)
# P(B)
mean(S==1)
# P(C)
mean(S==3)
# P(D)
1 - mean(S==0)
```

In the game of craps the "shooter" wins on the first roll of a pair of fair dice if the sum of the two dice is 7 or 11.  Calculate the probability that the shooter wins on the first roll.

Note there are 8 ways to get either a 7 or 11 on the first roll and 36 possible outcomes from rolling two dice.  So the $P(Win) = 8/36$.

Lets simulate rolling two dice 10,000 times.

```{r}
roll1 <- sample(1:6, 10000, replace = TRUE)
roll2 <- sample(1:6, 10000, replace = TRUE)
DF <- data.frame(roll1, roll2)
head(DF)
library(tidyverse)
DF %>% 
  mutate(SUM = roll1 + roll2) %>% 
  summarize(mean(SUM ==7 | SUM ==11))
```

