---
title: "In-Class Regression Work"
author: "Alan T. Arnholt"
date: 'Last compiled: `r format(Sys.time(), "%A, %B %d, %Y - %X.")`'
output: bookdown::html_document2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = NA, warning = FALSE, message = FALSE)
library(tidyverse)
library(PASWR2)
library(moderndive)
```

# Apartments in Vitoria

Making a new factor variable for elevators named `elevatorF`.

```{r}
VIT2005 <- VIT2005 %>% 
  mutate(elevatorF = as.factor(elevator)) %>% 
  glimpse()
```

## Exercise

* Create a scatterplot of `totalprice` versus `area`.

```{r, label = "vit", fig.cap = "No this is Boo playing a joke"}
ggplot(data = VIT2005, aes(x = area, y = totalprice)) +
  geom_point() + 
  theme_bw() +
  labs(x = "Area in square meters", y = "Price in Euros")
```

To refer to the Figure \@ref(fig:vit), use the syntax `\@ref(fig:label)`.

* Compute $r$ using the formula given in \@ref(eq:pearson) and verify your answer using the built-in R function `cor()`.

\begin{equation}
r = \frac{1}{n-1}\sum_{i=1}^{n}\left(\frac{x_i - \bar{x}}{s_x}\right)\cdot\left(\frac{y_i - \bar{y}}{s_y}\right)
(\#eq:pearson)
\end{equation}

```{r}
values <- VIT2005 %>% 
  mutate(x_xbar = area - mean(area), y_ybar = totalprice - mean(totalprice),
         zx = x_xbar/sd(area), zy  = y_ybar/sd(totalprice)) %>% 
  select(area, x_xbar, zx, totalprice, y_ybar, zy)
head(values)
values %>% 
  summarize(r = sum(zx*zy)/(length(zx) - 1))
#
#  Built In Function
VIT2005 %>% 
  summarize(r = cor(totalprice, area))
```


* Create a linear model object named `mod1` by regressing 
`totalprice` on to `area`.

```{r}
mod1 <- lm(totalprice ~ area, data = VIT2005)
mod1
```

* Write the least squares regression equation.

$$\widehat{\text{totalprice}}  = `r coef(mod1)[1]` + `r coef(mod1)[2]`\cdot \text{area}$$

* Use the coefficients from `mod1` and `mutate()` to compute the $\hat{y}$ and $e_i$ values.

```{r}
values <- VIT2005 %>%
  mutate(yhat = coef(mod1)[1] + coef(mod1)[2]*area,
         e = totalprice - yhat) %>% 
  select(totalprice, yhat, area, e)
head(values)
# Or
values <- VIT2005 %>% 
  mutate(yhat = predict(mod1), e = resid(mod1)) %>% 
  select(totalprice, yhat, area, e)
head(values)

```
* Find the least squares coeffients for regressing `totalprice` on to `area` using the formulas below.  

$$b_1 = r\cdot\frac{s_y}{s_x}$$
$$b_0 = \bar{y} - b_1\cdot\bar{x}$$


```{r}
VIT2005 %>% 
  summarize(r = cor(totalprice, area), 
            b1 = r*sd(totalprice)/sd(area), 
            b0 = mean(totalprice) - b1*mean(area))
```

* Print the least squares coefficients for `mod1` using `get_regression_table()`.

```{r}
CT <- get_regression_table(mod1)
CT
CT[1, 2]
CT[2, 2]
```

## Exercise

* Create the linear model object `mod2` by 
regressing `totalprice` on to `elevatorF`.

```{r}

```


* What `totalprice` do you predict for an apartment without an elevator and with an elevator (use the output from `mod2` only).

```{r}

```


* Answer the previous question using `group_by()` and `mean`.

```{r}

```


## Exercise

* Use the `evals` data set to create a parallel slopes model by regressing `score` on `bty_avg` and `rank`.  Store the result in `mod4`.

```{r}
mod4 <- lm(score ~ bty_avg + rank, data = evals)
mod4
```

* Write out the least squares regression lines for the three ranks.



* Graph `mod4` with `ggplot` (parallel slopes model).

```{r}

```

________________

# Plotly and Directories

## Creating a Directory

```{r}
url <- "http://faculty.marshall.usc.edu/gareth-james/ISL/Advertising.csv"

if(!dir.exists("../Data/")){
dir.create("../Data/")
}

if(!file.exists("../Data/Advertising.csv")){
    download.file(url, destfile = "../Data/Advertising.csv")}
Ad <- read.csv("../Data/Advertising.csv")
head(Ad)
# Want to create a 3D scatterplot of sales ~ TV + Radio
library(plotly)
p <- plot_ly(data = Ad, z = ~sales, x = ~TV, y = ~radio, opacity = 0.6) %>%
  add_markers(marker = list(size = 5)) 
p
```

Next we want to add the best fit plane to the 3D scatterplot.  Consider the best fit plane `sales_mod`:

```{r}
sales_mod <- lm(sales ~TV + radio, data = Ad)
summary(sales_mod)
```

```{r}
x <- seq(0, 300, length = 70)
y <- seq(0, 50, length = 70)
plane <- outer(x, y, function(a, b){summary(sales_mod)$coef[1,1] + 
    summary(sales_mod)$coef[2,1]*a + summary(sales_mod)$coef[3,1]*b})
# draw the plane
p %>%
  add_surface(x = ~x, y = ~y, z = ~plane, showscale = FALSE)
```

Consider a model with interaction:

```{r}
int_mod <- lm(sales ~ TV + radio + TV:radio, data = Ad)
summary(int_mod)
```

```{r}
plane_int <- outer(x, y, function(a, b){summary(int_mod)$coef[1,1] + 
    summary(int_mod)$coef[2,1]*a + summary(int_mod)$coef[3,1]*b +
    summary(int_mod)$coef[4,1]*a*b})
p %>% 
  add_surface(x = ~x, y = ~y, z = ~plane_int, showscale = FALSE)
```

```{r}
library(ISLR)
head(Credit)
mod <- lm(Balance ~ Limit + Income, data = Credit)
summary(mod)
p <- plot_ly(data = Credit, z = ~Balance, x = ~Limit, y = ~Income, opacity = 0.6) %>%
  add_markers(marker = list(size = 5)) 
p
```

```{r}
x <- seq(855, 13913, length = 70)
y <- seq(10, 187, length = 70)
plane <- outer(x, y, function(a, b){summary(mod)$coef[1,1] + 
    summary(mod)$coef[2,1]*a + summary(mod)$coef[3,1]*b})
# draw the plane
p %>%
  add_surface(x = ~x, y = ~y, z = ~plane, showscale = FALSE)
```

