---
title: "Chi-Square tests of Independence"
format: html
editor: visual
---
## Cleaning up the data

```{r, message = FALSE, warning = FALSE}
library(tidyverse)
library(infer)
load("../Data/gss.RData")
gss2016 <- gss |>
  filter(year == 2016) |> 
  # I am filtering to get a similar set used in DataCamp
  filter(!is.na(consci)) |> 
  filter(!is.na(cappun)) |> 
  filter(!is.na(sex)) |> 
  filter(!is.na(partyid)) |> 
  filter(!is.na(postlife)) |> 
  filter(!is.na(natspac)) |> 
  filter(!is.na(natarms)) |> 
  filter(!is.na(happy)) |> 
  filter(!is.na(region))
# Recoding levels of consci---similar to data camp???
gss2016$consci <- ifelse(gss2016$consci == "A GREAT DEAL", "High", "Low")
set.seed(321)  # Make same size as data camp...randomly
ind <- sample(1:nrow(gss2016), 150, replace = FALSE)
gss2016 <- gss2016[ind, ]
# Create new variable party
gss2016$party <- ifelse(gss2016$partyid %in% c("STRONG DEMOCRAT",
                                               "NOT STR DEMOCRAT"),
                        "Dem", ifelse(gss2016$partyid %in% 
                          c("IND,NEAR DEM", "IND,NEAR REP",
                            "INDEPENDENT"), "Ind",
                          ifelse(gss2016$partyid %in% 
                                   c("STRONG REPUBLICAN", 
                                     "NOT STR REPUBLICAN"), 
                                 "Rep", "Oth")))
###
# Subset data
gss_party <- gss2016 |>
  # Filter out the "Oth"
  filter(party != "Oth") 
tab_politics <- gss_party |>
  # Select columns of interest
  select(natarms, party) |>
  # Create table
  table()
# View the result
tab_politics
```
## Test of Independence Hypotheses

We would like to test the null hypothesis that opinions on how much money is spend on defense (`natarms`) is independent of party (`party`) affiliation.  The alternative hypothesis is that there is an association between opinions on how much money is spend on defense and party affiliation.

## Graphing the Counts

```{r}
gss_party |> 
  ggplot(aes(x = party, fill = natarms)) +
  geom_bar() + 
  theme_bw()
```

If we want to look at the proportions we would code the graph as follows:

```{r}
gss_party |> 
  ggplot(aes(x = party, fill = natarms)) +
  geom_bar(position = "fill") + 
  theme_bw()
```

## Computing the expected counts

```{r}
tab_politics -> obs
obs
chisq.test(obs)$expected -> exp
exp
```

## Computing the test statistic

```{r}
chi_sq_stat <- sum((obs - exp)^2/exp)
chi_sq_stat
```

## Computing the p-value

```{r}
pchisq(chi_sq_stat, 4, lower = FALSE)
# Making assumptions and using chisq.test()
chisq.test(obs, correct = FALSE)
```

## Verifying the result without assumptions

```{r}
null_natarms <- gss_party |> 
  specify(natarms ~ party) |> 
  hypothesize(null = "independence") |> 
  generate(reps = 10000, type = "permute") |> 
  calculate(stat = "Chisq")
visualize(null_natarms) + 
  shade_p_value(chi_sq_stat, direction = "right")
get_p_value(null_natarms, obs_stat = chi_sq_stat, direction = "right")
```

## Doing the same resampling process with a for loop

```{r}
set.seed(31)
(T1 <- table(gss_party$natarms, gss_party$party))
(obs_stat <- chisq.test(T1, correct = FALSE)$stat)
P <- 10^4
chi_stat <- numeric(P)
for(i in 1:P){
  chi_stat[i] <- chisq.test(table(gss_party$natarms, sample(gss_party$party)), correct = FALSE)$stat
}
hist(chi_stat)
(p_value <- mean(chi_stat >= obs_stat))
```
## Conclusion

Given the p-value ($`r p_value`$) is significantly less than $\alpha = 0.05$ we reject the null hypothesis and conclude there is an **association** between a persons opinion on the how much the country is spending on national defense and their party affiliation.
