---
title: "Linear Algebra Practice"
format: pdf
editor: source
---

1.  Use the following code to read in data and store the results in `Credit`

```{r}
Credit <- read.csv("http://statlearning.com/s/Credit.csv")
```

2.  Create a linear model object named `mod_num` by regressing `Balance` onto `Income`, `Limit`, `Rating`, `Cards` and `Age`.

```{r}
mod_num <- lm(Balance ~ Income + Limit + Rating + Cards + Age, data = Credit)
mod_num
summary(mod_num)
```

3.  Create the model matrix and store the results in `X`.

```{r}
X <- model.matrix(mod_num)
head(X)
```

4.  Create the matrix $X'X$ and store the result in `XTX`.

```{r}
XTX <- t(X)%*%X
XTX
```

\pagebreak

5.  Find the inverse of $(X'X)^{-1}$ and store the result in `XTXI`.

```{r}
XTXI <- solve(XTX)
XTXI
```


6.  Verify that the answer you computed in 5. is the same thing as `summary(mod_num)$cov.unscaled`.

```{r}
summary(mod_num)$cov.unscaled -> p6
p6
all.equal(XTXI, p6, tolerance = 0.0000001)
```


7.  Use the normal equations to find the coefficients for `mod_num`. Verify that your estimates computed with the normal equations are the same as `coef(mod_num)`.

```{r}
betahat <- XTXI%*%t(X)%*%Credit$Balance
betahat
coef(mod_num)
```

8.  What is the MSE for `mod_num`.

```{r}
MSE <- summary(mod_num)$sigma^2
MSE
```


9.  Compute the variance covariance matrix for your estimated coefficients. Store the values in an object named `var_cov_b`.

```{r}
var_cov_b <- MSE*XTXI
var_cov_b
```


10. Compute the standard errors for your estimated coefficients. Verify that your answers are the same as `summary(mod_num)$coef[, 2]`.

```{r}
se_b <- diag(var_cov_b)^.5
se_b
summary(mod_num)$coef[,2]
```


11. Compute the t values using your answers from 7. and 10. Verify that the values are the same as `summary(mod_num)$coef[, 3]`

```{r}
Ts <- betahat/se_b
Ts
summary(mod_num)$coef[, 3]
```


12. Compute the p-values for testing $H_0:\beta_{i} = 0$ versus $H_0:\beta_{i}\neq 0$ for $i =0,...,5$.

```{r}
ps <- pt(abs(Ts), 394, lower = FALSE)*2
ps
summary(mod_num)$coef[, 4]
```

